---
title: "Project Report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Project Report}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

*Note:* Before running the code please generate the
merged_data_cleaned.csv. To do this ensure you have saved the raw data
files from the repository and then run Cleaning And Merging The Data.R
to generate the merged_data_cleaned.csv. Due to file size limitations we
were unable to upload the compiled merged_data_cleaned.csv file
directly.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(lubridate)
library(ggplot2)
library(gratia)
library(mgcv)
library(caret)
library(DescTools)
library(scales)
library(tidyr)

# Load merged data
merged_data <- read.csv("../merged_data_cleaned.csv")
merged_df <-merged_data
##Research Question #1

# Replace rain NA with 0 (NA indicates no rain was recorded)
merged_df$rain[is.na(merged_df$rain)] <- 0

# Remove rows with missing price
merged_df <- merged_df %>% filter(!is.na(price))

# Create useful derived variables
# Data was collected in Boston (EST = UTC-5 in Nov-Dec 2018)
merged_df <- merged_df %>%
  mutate(
    time_stamp  = as.POSIXct(time_stamp, tz = "UTC"),
    hour_ts     = as.POSIXct(hour_ts, tz = "UTC"),
    time_local  = with_tz(time_stamp, tzone = "America/New_York"),
    hour_of_day = hour(time_local),
    day_of_week = lubridate::wday(time_local, label = TRUE),
    cab_label   = paste(cab_type, name, sep = "-"),
    route       = paste(source, "->", destination),
    surge_binary = ifelse(surge_multiplier > 1, 1, 0)
  )

cat("Dataset dimensions:", nrow(merged_df), "rows x", ncol(merged_df), "columns\n")
cat("Date range:", as.character(min(merged_df$time_stamp)),
    "to", as.character(max(merged_df$time_stamp)), "\n")


##Research Question #2
# Handle missing values
merged_df$rain[is.na(merged_df$rain)] <- 0
merged_df <- merged_df %>% filter(!is.na(price))

# Create derived variables
# Convert to Boston local time (EST = UTC-5 in Nov-Dec 2018) before extracting hour/day
merged_df <- merged_df %>%
  mutate(
    time_stamp   = as.POSIXct(time_stamp, tz = "UTC"),
    time_local   = with_tz(time_stamp, tzone = "America/New_York"),
    hour_of_day  = hour(time_local),
    # Bin hours into meaningful time periods to capture non-linear hourly patterns
    time_period  = factor(case_when(
      hour_of_day >= 6  & hour_of_day < 10 ~ "morning_rush",
      hour_of_day >= 10 & hour_of_day < 16 ~ "midday",
      hour_of_day >= 16 & hour_of_day < 20 ~ "evening_rush",
      TRUE                                  ~ "night"
    ), levels = c("midday", "morning_rush", "evening_rush", "night")),
    # Use unordered factor so R applies treatment (dummy) coding, not polynomial contrasts
    day_of_week  = factor(lubridate::wday(time_local, label = TRUE), ordered = FALSE),
    route        = paste(source, "->", destination),
    surge_binary = ifelse(surge_multiplier > 1, 1, 0)
  )


##Research Question #3
merged_data <- merged_data %>%
  mutate(
    hour_ts_boston = with_tz(ymd_hms(hour_ts, tz = "UTC"), tzone = "America/New_York"),
    date = as.Date(hour_ts_boston),
    hour = hour(hour_ts_boston),
    day_of_week = factor(wday(date, label = TRUE), ordered = FALSE),
    is_weekend = ifelse(wday(date) %in% c(1, 7), 1,0),
    rush_hour = case_when(
      is.na(hour) ~ NA_character_, #Some data points don't have a time
      hour >= 6 & hour <= 10 ~ "Morning Rush",
      hour >= 16 & hour <= 20 ~ "Evening Rush",
      TRUE ~ "Normal Hours"
    ),
    holiday = case_when(
      date == as.Date("2018-11-26") ~ "Cyber Monday",
      date >= as.Date("2018-12-02") & date <= as.Date("2018-12-10") ~ "Hanukkah",
      TRUE ~ "None"
    ),
    cab_combo = paste(cab_type, name, sep = "-")
  )

merged_data <- merged_data %>% filter(!is.na(hour))

lyft_data <- merged_data %>%
  filter(cab_type == "Lyft")

raw <- read.csv("../merged_data_cleaned.csv") #Shows how many data points don't have a time
sum(!grepl(":", raw$hour_ts_boston))

class(merged_data$hour_ts_boston)
head(merged_data$hour_ts_boston)
```

**Front Matter** 

Uber and Lyft Surge Pricing Analysis 
By Lingyun Jiang (B00850695) and Sydney Parsons (B01068132) 
Data Analysis STAT 4620/5620 Winter 25-26 
Link to GitHub:
<https://github.com/SydneyEParsons/Uber_Lyft_SurgePricing_Analyis>

## Abstract (150 words)

This is an analysis of Uber and Lyft Pricing in Boston Massachusetts for 23 days in late November to early December of 2018 (Munde, 2019). In this paper we cover three main research questions with the goal of determining the overall impacts on pricing for Uber and Lyft. 

Ultimately we determined that the primary impacts on Uber and Lyft pricing are their cab types and their ride distance. The pricing for each of these ride shares are very segmented and stable within their type boundaries. Distance is the variable factor that changes the most, as riders choose rides that are further and shorter distances that will impact their ride cost. Although their prices do change with influences from our other variables mentioned, they are not significant enough when compared to the influence of cab type and distance on price.


##Keywords 

Uber, Lyft, Pricing, Surge, Holiday, Rushhour, DOW, Weather, Distance


##Introduction Research questions and background information.

**Research Question 1:**

**Research Question 2:**

**Research Question 3:**



## Data Description:

-   Data points description here\*\*


**Variables:**



### EDA: Research Question 1

**Figure 1: Distribution of Ride Prices**

```{r}
ggplot(merged_df, aes(x = price)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Ride Prices",
       x = "Price (USD)", y = "Count") +
  theme_minimal()
```

For our first research question we began by examining the distribution of ride prices. As shown in Figure 1, the histogram revealed a strong right skew with the majority of rides concentrated at lower prices and a long tail extending toward more expensive service tiers. This pattern suggested that a model assuming normally distributed errors would not be appropriate and that the Gamma distribution would better capture the shape of the response variable.

**Figure 2: Temperature vs. Price (Log Scale) by Platform:**

```{r}
set.seed(42)
sample_df <- merged_df %>% slice_sample(n = 20000)

ggplot(merged_df, aes(x = temp, y = price)) +
  geom_point(alpha = 0.02, size = 0.3, color = "grey40") +
  geom_smooth(data = sample_df, method = "gam", formula = y ~ s(x),
              se = TRUE, color = "steelblue", linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~ cab_type) +
  labs(title = "Temperature vs. Price (Log Scale) by Platform",
       x = "Temperature (F)", y = "Price (USD, log scale)") +
  theme_minimal()
```

We then plotted temperature against price on a log scale with a GAM smoother overlaid on a random subsample. As shown in Figure 2, the smoother revealed a slight non-linear trend, but the effect was visually minor compared to the spread of the data. This indicated that while a non-linear relationship may exist, its practical magnitude is small relative to other sources of price variation.

**Figure 3: Correlation Between Weather and Price by Service Type:**

```{r}
cor_results <- merged_df %>%
  group_by(cab_label) %>%
  summarise(
    Temperature = cor(price, temp,     use = "complete.obs"),
    Clouds      = cor(price, clouds,   use = "complete.obs"),
    Pressure    = cor(price, pressure, use = "complete.obs"),
    Humidity    = cor(price, humidity,  use = "complete.obs"),
    Wind        = cor(price, wind,     use = "complete.obs"),
    Rain        = cor(price, rain,     use = "complete.obs"),
    .groups = "drop"
  )

cor_long <- cor_results %>%
  pivot_longer(-cab_label, names_to = "Weather Variable", values_to = "Correlation")

ggplot(cor_long, aes(x = `Weather Variable`, y = cab_label, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 3)), size = 2.8) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  labs(title = "Correlation Between Weather and Price by Service Type",
       subtitle = "Unadjusted Pearson correlations (not controlling for distance or route)",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Figure 3 presents the unadjusted Pearson correlations between weather variables and price for each service type. The correlations were uniformly weak across all service types with most values falling between negative 0.05 and positive 0.05. Temperature showed the strongest individual correlation, but it remained small in absolute terms. These patterns indicated that weather variables are unlikely to be dominant price drivers, motivating our use of a model that can detect subtle and potentially non-linear effects.

### EDA: Research Question 2
```{r}
# Confirm Uber surge is always 1, only Lyft has variation
merged_df %>%
  group_by(cab_type) %>%
  summarise(
    n_rides      = n(),
    unique_surge = n_distinct(surge_multiplier),
    min_surge    = min(surge_multiplier),
    max_surge    = max(surge_multiplier),
    pct_surge    = round(mean(surge_multiplier > 1) * 100, 2),
    .groups = "drop"
  )
```

For our second research question we focused on the Lyft subset of the data because Uber showed a constant surge multiplier of 1.0 throughout the entire dataset. Shared rides were excluded as they also showed no surge variation, which would cause estimation problems in a logistic regression.

```{r}
# Focus on Lyft only, exclude Shared rides
lyft_df <- merged_df %>%
  filter(cab_type == "Lyft", name != "Shared")

cat("Lyft rides (excl. Shared):", nrow(lyft_df), "\n")
cat("Surge rate:", round(mean(lyft_df$surge_binary) * 100, 2), "%\n\n")
cat("Surge multiplier distribution:\n")
table(lyft_df$surge_multiplier)
```

**Figure 1: Surge Rate by Pickup Location:**

```{r}
source_surge <- lyft_df %>%
  group_by(source) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(source_surge, aes(x = reorder(source, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Pickup Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

As shown in Figure 1, there was substantial variation in surge rates by pickup location with some locations showing surge rates above 30 percent while others remained below 10 percent. This suggested that geographic factors play a significant role in surge occurrence.

**Figure 2: Surge Rate by Dropoff Location:**

```{r}
dest_surge <- lyft_df %>%
  group_by(destination) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(dest_surge, aes(x = reorder(destination, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Dropoff Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

Figure 2 presents a similar analysis for dropoff locations. The variation across destinations was also present though somewhat less pronounced than for pickup locations, indicating that where a ride begins may matter more than where it ends for surge pricing.

**Figure 3: Hourly Surge Rate (Lyft):**

```{r}
hour_surge <- lyft_df %>%
  group_by(hour_of_day) %>%
  summarise(surge_rate = mean(surge_binary), n_rides = n(), .groups = "drop")

ggplot(hour_surge, aes(x = hour_of_day, y = surge_rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Hourly Surge Rate (Lyft)",
       x = "Hour of Day", y = "Surge Rate") +
  theme_minimal()
```

Figure 3 showed that surge rates peaked during morning and evening commute hours, though the magnitude of the difference was moderate. The highest surge rates were observed in the early morning hours while midday showed the lowest rates overall.

**Figure 4: Weather Conditions: Surge vs. Non-Surge:**

```{r}
weather_vars <- c("temp", "wind", "humidity", "clouds", "pressure", "rain")

# Standardize weather variables for comparable scale
lyft_std <- lyft_df %>%
  mutate(across(all_of(weather_vars), ~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE))) %>%
  mutate(Surge = ifelse(surge_binary == 1, "Surge", "No Surge")) %>%
  select(Surge, all_of(weather_vars)) %>%
  pivot_longer(-Surge, names_to = "Variable", values_to = "Value")

lyft_means <- lyft_std %>%
  group_by(Surge, Variable) %>%
  summarise(mean_val = mean(Value, na.rm = TRUE), .groups = "drop")

ggplot(lyft_means, aes(x = Variable, y = mean_val, fill = Surge)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7),
           width = 0.6, alpha = 0.8) +
  scale_fill_manual(values = c("No Surge" = "grey60", "Surge" = "steelblue")) +
  labs(title = "Weather Conditions: Surge vs. Non-Surge (Lyft)",
       subtitle = "Standardized means (z-scores); small differences suggest weather is not the primary surge driver",
       x = NULL, y = "Standardized Mean") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

Figure 4 compared standardized weather variable means between surge and non-surge rides. The differences were extremely small across all weather variables, providing early evidence that weather is not the primary trigger for surge pricing.

**Figure 5: Surge Rate by Time Period and Day of Week:**

```{r}
time_day_surge <- lyft_df %>%
  group_by(time_period, day_of_week) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  )

ggplot(time_day_surge, aes(x = day_of_week, y = time_period, fill = surge_rate)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = percent) +
  labs(title = "Surge Rate by Time Period and Day of Week (Lyft)",
       x = NULL, y = NULL, fill = "Surge Rate") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Figure 5 presents a heatmap of surge rates across time periods and days of the week. The heatmap confirmed that temporal patterns exist but are secondary to location effects. Certain combinations of time period and day showed slightly elevated surge rates, but no single temporal factor dominated the way that pickup location did.

### EDA: Research Question 3

**Figure 1: Price Distribution By Cab Type:**

```{r}
ggplot(merged_data, aes(x = cab_combo, y = price, fill = cab_combo)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~ cab_type, scales = "free_x") +
  scale_y_continuous(limits = c(0, 60)) +
  labs(title = "Price Distribution by Cab Type",
       x = "Cab Type", y = "Price") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```
To begin our EDA plots for our third research question we plotted our cab combo data compared to price. This was a starting point as we further breakdown the pricing analysis in the coming EDA plots. In this EDA plot we used boxplots to demonstrate this distribution and median of the pricing data per each category. As you can see from the visualization the plots show very segmented data. Both ride hailing apps have cab-types that are very regimented with their pricing, there is limited overlap between the pricing of different cab types. 

**Figure 2: Analyze Hourly Price on Normal Days of the Week:**

```{r}
#Analyze hourly price on normal days of the week
normal_rush_dayofweek_summary <- merged_data %>%
    filter(holiday=="None") %>%
    group_by(hour, day_of_week, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )

ggplot(data=normal_rush_dayofweek_summary %>% filter(cab_type == "Lyft"), mapping =aes(x=hour, y=mean_price, color=day_of_week)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Analyze hourly price on normal days of the week
", x="hour", y="mean_price") 

ggplot(data=normal_rush_dayofweek_summary %>% filter(cab_type == "Uber"), mapping =aes(x=hour, y=mean_price, color=day_of_week)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Analyze Hourly Price on Normal Days of the Week
", x="Hour", y="Mean Price") 
```
These line graphs show the mean pricing for the different cab combos during the days of the week. The lines overall remain very stable but there is greater variation with pricing in the higher price hailing tiers, specifically in Lyft-Lux Black, Lyft-Lux Black XL, Uber-Black, and Uber-Black SUV. These variances are not significant but it is clear there is more noise in these combos than the others.

**Figure 3: Compare Day of the Week, Price and Cab Combo:**

```{r}
#Compare day of week, price and cab_combo
day_of_week <- merged_data %>%
  group_by(day_of_week, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(day_of_week, n = Inf)

ggplot(data=day_of_week, mapping =aes(x=day_of_week, y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)), size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Compare day of week, price and cab_combo
", x="Day of Week", y="Mean Price") +
  coord_flip() 

ggplot(data=day_of_week %>% filter(cab_type == "Lyft"), 
    mapping = aes(x=day_of_week, y=cab_combo, fill=mean_price)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2) +
labs(title = "Lyft - Compare day of week, price and cab_combo
", x="Day of Week", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data=day_of_week %>% filter(cab_type == "Uber"), 
    mapping = aes(x=day_of_week, y=cab_combo, fill=mean_price)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2) +
labs(title = "Uber - Compare day of week, price and cab_combo
", x="Day of Week", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Here we are focusing on the influence of the days of the week on the pricing of ride hailing platforms Uber and Lyft. Our following EDA plots is a colour coded bar plot and heat maps. These plots help illustrate the relationship between the days of the week and our cab combo. As apparent through the plots the pricing remains very consistent across each day of the week and cab types. 

**Figure 4: Price Distribution by Day of The Week:**

```{r}
ggplot(merged_data, aes(x = day_of_week, y = price, fill = day_of_week)) +
  geom_violin(trim = FALSE) +  
  geom_boxplot(width = 0.1, outlier.shape = NA) +  
  facet_wrap(~ cab_type) +
  scale_y_continuous(limits = c(0, 60)) +  
  labs(title = "Price Distribution by Day of The Week",
       x = "Day of Week", y = "Price") +
  theme_minimal() +
  theme(legend.position = "none") 
```
Since more variation was expected in the data as highlighted in our hypothesis, we plotted an additional EDA plot as a violin plot. This plot helps show the quantity of rides at each price point through the density of the waves. In our presentation it was highlighted that they were very uniform and that we should investigate further. After analysing the code we realized the smoothing we applied to the pots was distorting the final display. After removing that we were able to see the density of the prices more clearly. However, this plot reaffirms the lack of variation in pricing throughout the days of the week as seen in our initial EDA plots above.

**Figure 5: Compare Weekday and Weekends to Price and Cab Combo:**

```{r}
#Compare weekday and weekends to price and cab_combo
weekend_summary <- merged_data %>%
  group_by(is_weekend, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(weekend_summary)

ggplot(data=weekend_summary, mapping =aes(x=factor(is_weekend, labels=c("Weekday", "Weekend")), y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)),  size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Compare weekday and weekends to price and cab_combo", x="Is Weekend", y="Mean Price") +
  coord_flip() 
```
With little variation in the days of the week we analysed the data further. By plotting the week day against the weekend we aimed to try to identify price variations on a more general basis but this plot also did not show any notable variation in pricing from weekday to weekend.

**Figure 6: Price Distribution By Holiday:**

```{r}
#Comparing weekend and weekday pricing to holiday pricing
holiday_summary_detailed <- merged_data %>%
  group_by(holiday, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(holiday_summary_detailed)

ggplot(merged_data, aes(x = holiday, y = price, fill = holiday)) +
  geom_violin(trim = FALSE) +  
  geom_boxplot(width = 0.1, outlier.shape = NA) +  
  facet_wrap(~ cab_type) +
  scale_y_continuous(limits = c(0, 60)) +  
  labs(title = "Price Distribution by Holiday",
       x = "Holiday", y = "Price") +
  theme_minimal() +
  theme(legend.position = "none") 

ggplot(data=holiday_summary_detailed, mapping =aes(x=factor(holiday, labels=c("Cyber Monday", "Hanukkah", "None")), y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)),  size = 2.5, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Comparing Holiday Pricing to Cab Combo", x="Holiday", y="Mean Price") +
  coord_flip() 
```
For price distribution by holiday, we presumed traffic would be higher on these days due to people travelling for the holiday and having time off. However, as seen in the violin plots and the bar plots below that was not the case. This may indicate that the pricing was changing slightly between cab-type segments on Lyft rides. However, Uber does remain very constant in its pricing regardless of holiday. The bar graph also shows the Lyft ride prices being more incremental than Uber ride pricing which remains very stable. 

**Figure 7: Comparing Cost of Rides on Holidays:**

```{r}
#Compare time of day price on normal days to time of day on holiday
normal_time_holiday_time_summary <- merged_data %>%
    filter(is_weekend==0) %>%
    group_by(hour, holiday, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )

ggplot(data=normal_time_holiday_time_summary %>% filter(cab_type == "Lyft"), mapping =aes(x=hour, y=mean_price, color=holiday)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Compare Cost of Rides on Holidays", x="hour", y="mean_price") 

ggplot(data=normal_time_holiday_time_summary %>% filter(cab_type == "Uber"), mapping =aes(x=hour, y=mean_price, color=holiday)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Compare Cost of Rides on Holidays", x="hour", y="mean_price")
```
To further visualise the pricing of Uber and Lyft on holidays we plotted them on line graphs. This provides an additional visualization outside of the violin plots. It is clear from these plots that the prices remain stable with small peaks later in the days presumably around the evening rush hour period.

**Figure 8: Comparing Rush Hour Price for Weekdays to Price on Normal Weekends:**

```{r}
#Comparing rush hour price for weekdays to price on weekends
rush_weekdays_normal_weekends_summary <- merged_data %>%
  group_by(rush_hour, is_weekend, cab_combo, cab_type) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(rush_weekdays_normal_weekends_summary)

ggplot(data=rush_weekdays_normal_weekends_summary, mapping =aes(x=rush_hour, y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)),  size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ is_weekend,
  labeller = labeller(is_weekend = c("0" = "Weekday", "1" = "Weekend"))) +
labs(title = "Comparing rush hour price for weekdays to price on normal weekends
", x="Rush Hour", y="Mean Price") +
  coord_flip() 

ggplot(data=rush_weekdays_normal_weekends_summary %>% filter(cab_type == "Lyft"),
    mapping = aes(x=rush_hour, y=cab_combo, fill=mean_price)) +
geom_tile() +
facet_wrap(~ is_weekend, labeller = labeller(is_weekend = c("0" = "Weekday", "1" = "Weekend"))) +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2.5) +
labs(title = "Lyft - Comparing rush hour price for weekdays to price on normal weekends
", x="Rush Hour", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data=rush_weekdays_normal_weekends_summary %>% filter(cab_type == "Uber"),
    mapping = aes(x=rush_hour, y=cab_combo, fill=mean_price)) +
geom_tile() +
facet_wrap(~ is_weekend, labeller = labeller(is_weekend = c("0" = "Weekday", "1" = "Weekend"))) +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2.5) +
labs(title = "Uber - Comparing rush hour price for weekdays to price on normal weekends
", x="Rush Hour", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
In this EDA section we plotted the comparison of rush hour to weekdays and weekends on a bar plot categorized by cab-type. This plot demonstrates that the prices for rides remain steady across normal hours, morning rush and evening rush on both evenings and weekends. There are many bars on this plot and they are smaller but the numbers on the bars further confirm the limited change in pricing across the varying rush hours on weekdays and weekends.

**Figure 9: Price Distribution by Rush Hour:**

```{r}
ggplot(merged_data, aes(x = rush_hour, y = price, fill = rush_hour)) +
  geom_violin(trim = FALSE) +  
  geom_boxplot(width = 0.1, outlier.shape = NA) +  
  facet_wrap(~ cab_type) +
  scale_y_continuous(limits = c(0, 60)) +  
  labs(title = "Price Distribution by Rush Hour",
       x = "Rush Hour", y = "Price") +
  theme_minimal() +
  theme(legend.position = "none") 
```
The following violin plot demonstrates the price distribution of Uber and Lyft apps during our three categorized rush hour periods. As can be seen in the plot the price remains very similar across all three rush hour periods. The wider sections of the violin plots in this EDA indicate that there are more rides at that price point. Both Uber and Lyft have very specific shapes that do not change as rush hour changes.

**Figure 10: Compare Cab Combo, Holiday and Uber & Lyft Surge Multiplier:**

```{r}
#Compare cab_combo, holiday and surge multiplier
holiday_surge_summary <- merged_data %>%
  group_by(holiday, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE),
    mean_surge = mean(surge_multiplier, na.rm = TRUE),
    median_surge = median(surge_multiplier, na.rm = TRUE)
  )

ggplot(data=holiday_surge_summary, mapping =aes(x=cab_combo, y=mean_surge, fill=holiday)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_surge, 2)),  size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Compare Cab Combo, Holiday & Surge Multiplier", x="Cab Type", y="Surge Pricing") +
  coord_flip() 
```
Surge multiplier was highlighted in our earlier research questions but we did also briefly plot it here to test its influence in our third research question. Uber remains constant because there was no observed surge price in this collected dataset. You can see the surge prices for Uber and Lyft here but a bar plot was not the most ideal for comparing surge across cab combo and holiday which is why we pivoted to a heatmap as can be seen in the following plot.

**Figure 11: Lyft and Uber Compare Cab Combo, Holiday and Surge Multiplier:**

```{r}
ggplot(data=holiday_surge_summary %>% filter(cab_type == "Lyft"), 
    mapping = aes(x=holiday, y=cab_combo, fill=mean_surge)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_surge, 2))) +
labs(title = "Lyft - Compare Cab Combo, Holiday and Surge Multiplier
", x="Holiday", y="Cab Combo") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data=holiday_surge_summary %>% filter(cab_type == "Uber"), 
    mapping = aes(x=holiday, y=cab_combo, fill=mean_surge)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_surge, 2))) +
labs(title = "Uber - Compare Cab Combo, Holiday and Surge Multiplier
", x="Holiday", y="Cab Combo") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Here you can see a more clear breakdown of our surge visualization. A heatmap for this data is more appropriate because it shows the distribution through the colours and numbers on the plot in a much cleaner presentation. As we can see from the values of the surge multiplier for Lyft surge doesn’t change significantly over holidays compared to non-holidays.

**Figure 12: Compare Pricing on Weekday Holidays to Rush Hour Pricing:**

```{r}
#Compare pricing on weekday holidays to rush hour pricing
weekday_holiday_rush_summary <- merged_data %>%
    filter(is_weekend==0) %>%
    group_by(rush_hour, holiday, cab_type) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )

ggplot(data=weekday_holiday_rush_summary, mapping =aes(x=rush_hour, y=mean_price, fill=holiday)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
facet_wrap(~ cab_type) +
geom_text(aes(label=round(mean_price, 2)),  size = 2, position=position_dodge(width=0.9)) +
labs(title = "Compare Pricing on Weekday Holidays to Rush Hour Pricing") +
  coord_flip()

ggplot(data=weekday_holiday_rush_summary, mapping =aes(x=rush_hour, y=holiday, fill=mean_price)) +
geom_tile() +
facet_wrap(~ cab_type) +
  scale_fill_gradient(low="white", high="red") +
geom_text(aes(label=round(mean_price, 2))) +
labs(title = "Compare pricing on weekday holidays to rush hour pricing
", x="Rush Hour", y="Holiday") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
For our comparison of weekday holidays to rush hour we initially plotted a bar graph and then pivoted to a heatmap to better visualize the values. The prices over all Uber and Lyft rides during all three rush hour periods on holidays were very stable. There was a slight dip on the price of Lyft rides in the evenings of Hannukah, and a slight increase in price of Uber rides in the normal hours of Cyber Monday.

After completing our exploratory analyses we proceeded to create the models for each of our research questions. The information provided from our analysis helped us choose methods in an informed manner.

## Methods

**Research Question 1:** 
To assess the effect of weather on ride price we used a Generalized Additive Model with a Gamma family and log link function. This model was chosen based on three characteristics of the response variable.

First, ride prices are strictly positive, which the Gamma distribution naturally accommodates. Second, the distribution of prices is right skewed with a concentration of rides at lower prices and a long tail of expensive rides from premium service tiers. The Gamma distribution handles this asymmetry well. Third, more expensive rides exhibit greater price variability, which is consistent with the Gamma assumption that variance is proportional to the squared mean.

The log link ensures that predicted prices remain positive and that covariate effects act multiplicatively on the price scale. This is a natural interpretation for pricing data because a one unit change in temperature, for example, would scale the price by a constant percentage rather than adding a fixed dollar amount.

We used smooth functions for the six weather variables to allow for potentially non-linear relationships, while distance, cab type, service name, source, and destination entered as parametric terms. The basis dimensions were set to reflect the expected complexity of each relationship. Temperature was assigned k equals 8 because it had the widest range and the most potential for complex non-linearity. Clouds and pressure were assigned k equals 5 for moderate flexibility. Wind, humidity, and rain were each assigned k equals 4 because the data showed limited variability in these variables or, in the case of rain, very sparse observations. The effective degrees of freedom reported in the model output indicate how much of this flexibility the data actually require.

The model was fit using the bam function from the mgcv package (Wood, 2017), which is a computationally efficient version of gam designed for large datasets. We used fast REML for smoothing parameter estimation with the discrete option enabled to further accelerate computation.

The key assumptions of this model are that conditional on covariates, prices follow a Gamma distribution with a constant shape parameter, and that the smooth terms can adequately approximate the true underlying relationships. We also assume that observations are conditionally independent given the covariates. We acknowledge that repeated sampling of the same routes over time may introduce correlation, which is discussed further in the Limitations section.

**Research Question 2:** 
To model factors associated with surge pricing we used a logistic regression with a binomial family and logit link, treating surge occurrence as a binary outcome. A ride was classified as having surge if the surge multiplier exceeded 1.0 and as not having surge otherwise. This binary approach was chosen because the surge multiplier in the Lyft data takes only a few discrete values, with the vast majority of observations at 1.0 or 1.25. The extreme rarity of higher surge levels made a continuous or ordinal model impractical.

The model includes weather variables consisting of temperature, wind, humidity, cloud cover, pressure, and rain, as well as temporal variables including time period and day of week, pickup location, dropoff location, and service type as fixed effects. The time period was categorised into four bins: morning rush from 6am to 10am, midday from 10am to 4pm, evening rush from 4pm to 8pm, and night from 8pm to 6am.

As a robustness check we also fit a Generalized Linear Mixed Model with random intercepts for source and destination locations using the lme4 package (Bates et al., 2015). The purpose of this GLMM was to assess whether treating locations as random effects, which partially pools location level estimates and accounts for potential within location correlation, would change our conclusions about the weather and time predictors.

It is important to note that we do not use AIC or likelihood ratio tests to compare the GLM and GLMM directly. These two models use different likelihood formulations. The GLM maximises a conditional likelihood while the GLMM with Laplace approximation uses an approximate marginal likelihood that integrates over random effects. As discussed in the course materials, standard AIC is only valid for comparing GLMM variants when the goal is to account for correlation structures. Comparing across model classes requires conditional AIC, which is beyond the scope of this analysis. Instead, we evaluate the necessity of random effects by examining the estimated random effect variances, the consistency of fixed effect estimates between models, and the practical context of our grouping structure.

The assumptions of the logistic GLM include independent observations conditional on covariates and a linear relationship between covariates and the log odds of surge. The GLMM additionally assumes that random effects are normally distributed. Independence is approximate given the repeated measures nature of the data, but with large sample sizes per group this is unlikely to affect the direction of our conclusions.

**Research Question 3:** 
For research question three where we analysed different effects on cab prices we chose to use a GLM. Initially we had decided to compare a GLM and a GAM to see which one would perform best on the dataset. 

For the GAM, price was our response variable and the covariates used were: holiday, day of the week, hour, cab combo and surge multiplier. For our research question these were the covariates from our data that represented our model the best. For the covariate of hour we applied a smoothing function since it is continuous. The model distribution we used was Gamma with a log link. This was chosen to best fit the data of the response variable because it is positive and continuous, and the link function for Gamma distributions is the log link function this keeps the values positive (Kurui, 2025).
For the GLM to remain consistent for our comparison price was also used as our response variable and maintained the same covariates. Gamma distribution was chosen for the response variable because price is positive and continuous and the log link function was also used again to ensure the values remained positive as required under a Gamma distribution (Kurui, 2025).  

Assumptions of our GLM and GAM models include price being continuous, the data correctly represents the Uber and Lyft prices in Boston Massachusetts, relationship is assumed to be linear, distribution is positive continuous, measurements and observations are independent (Harrell, 2024). However we did have differing assumptions on the behaviour of the data as related to the GLM and GAM models themselves. In our GAM we assumed the data would present non-linearly, and with our GLM we assumed the data would present linearly.

To evaluate the models we plotted the residuals of both the GAM and GLM to visualize how they fit the data and the variance and distribution of the data points. We used the pseudo r-squared McFadden values and k-fold cross validation to test the fit of our GLM. And for our GAM we used r-squared to assess deviance explained by the model and EDF to review the linearity of the model.


## Analysis

### Research Question #1 

```{r}
gam_model <- bam(
  price ~ s(temp, k = 8) + s(wind, k = 4) + s(humidity, k = 4) +
          s(clouds, k = 5) + s(pressure, k = 5) + s(rain, k = 4) +
          distance + cab_type + name + source + destination,
  data    = merged_df,
  family  = Gamma(link = "log"),
  method  = "fREML",
  discrete = TRUE
)
 
summary(gam_model)
```

```{r}
par(mfrow = c(3, 2), mar = c(4, 4, 2, 1))
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE)
par(mfrow = c(1, 1))
```

```{r}
par(mfrow = c(2, 2))
gam.check(gam_model)
par(mfrow = c(1, 1))
```

```{r}
cat("Adjusted R-squared:", summary(gam_model)$r.sq, "\n")
cat("Deviance explained:", round(summary(gam_model)$dev.expl * 100, 2), "%\n")
```

### Research Question #2 

```{r}
glm_surge <- glm(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week +
                 source + destination + name,
  data   = lyft_df,
  family = binomial(link = "logit")
)
 
summary(glm_surge)
```

```{r}
anova(glm_surge, test = "Chisq")
```

```{r}
glmm_surge <- glmer(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week + name +
                 (1 | source) + (1 | destination),
  data    = lyft_df,
  family  = binomial(link = "logit"),
  nAGQ    = 0,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000))
)
 
summary(glmm_surge)
```

```{r}
# 1. Check random effect variance estimates
cat("=== Random Effect Variance Estimates ===\n")
vc <- VarCorr(glmm_surge)
print(vc)
 
# 2. Compare fixed effect estimates between GLM and GLMM
cat("\n=== Fixed Effect Comparison (Weather & Time) ===\n")
glm_coefs  <- coef(glm_surge)
glmm_coefs <- fixef(glmm_surge)
 
common_vars <- intersect(names(glm_coefs), names(glmm_coefs))
weather_time_pattern <- c("temp", "wind", "humidity", "clouds", "pressure",
                          "rain", "time_period", "day_of_week")
keep <- sapply(common_vars, function(x)
  any(sapply(weather_time_pattern, function(p) startsWith(x, p))))
compare_vars <- common_vars[keep]
 
comparison_df <- data.frame(
  Variable   = compare_vars,
  GLM_coef   = round(glm_coefs[compare_vars], 5),
  GLMM_coef  = round(glmm_coefs[compare_vars], 5),
  Difference = round(glm_coefs[compare_vars] - glmm_coefs[compare_vars], 5)
)
print(comparison_df, row.names = FALSE)
 
# 3. Contextual argument
cat("\n=== Contextual Justification ===\n")
cat("Number of source locations:", length(unique(lyft_df$source)), "\n")
cat("Number of destination locations:", length(unique(lyft_df$destination)), "\n")
cat("Average rides per source:", round(nrow(lyft_df) / length(unique(lyft_df$source))), "\n")
cat("Average rides per destination:", round(nrow(lyft_df) / length(unique(lyft_df$destination))), "\n")
```

```{r}
# McFadden pseudo R-squared for the logistic GLM
cat("McFadden Pseudo R-squared:", PseudoR2(glm_surge, which = "McFadden"), "\n")
```

```{r}
fe <- coef(glm_surge)
se <- summary(glm_surge)$coefficients[, "Std. Error"]
 
keep_pattern <- c("temp", "wind", "humidity", "clouds", "pressure", "rain", "time_period")
idx <- sapply(names(fe), function(x) any(sapply(keep_pattern, function(p) startsWith(x, p))))
 
or_df <- data.frame(
  Variable = names(fe)[idx],
  OR       = exp(fe[idx]),
  Lower    = exp(fe[idx] - 1.96 * se[idx]),
  Upper    = exp(fe[idx] + 1.96 * se[idx])
)
or_df$Variable <- gsub("time_period", "Time: ", or_df$Variable)
 
ggplot(or_df, aes(x = reorder(Variable, OR), y = OR)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title = "Odds Ratios for Surge Occurrence (GLM)",
       x = NULL, y = "Odds Ratio (95% CI)") +
  theme_minimal()
```

```{r}
fitted_vals <- fitted(glm_surge)
resids      <- residuals(glm_surge, type = "response")
 
n_bins <- 40
bin_breaks <- quantile(fitted_vals, probs = seq(0, 1, length.out = n_bins + 1))
bins <- cut(fitted_vals, breaks = bin_breaks, include.lowest = TRUE)
 
bin_df <- data.frame(fitted_vals, resids, bins) %>%
  group_by(bins) %>%
  summarise(
    avg_fitted = mean(fitted_vals),
    avg_resid  = mean(resids),
    se_resid   = sd(resids) / sqrt(n()),
    n          = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(bins))
 
ggplot(bin_df, aes(x = avg_fitted, y = avg_resid)) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = -2 * se_resid, ymax = 2 * se_resid),
              alpha = 0.15, fill = "grey50") +
  labs(title = "Binned Residual Plot (GLM)",
       x = "Average Fitted Probability", y = "Average Residual") +
  theme_minimal()
```

### Research Question #3 

```{r}
#Modelling GAM
gam_model_price <- gam(price ~ holiday + day_of_week + s(hour) + cab_combo + surge_multiplier,
                      data = merged_data,
                      family = Gamma(link = "log"))
summary(gam_model_price)
appraise(gam_model_price)
draw(gam_model_price)
```

```{r}
#Modelling GLM
glm_model_price <- glm(price ~ holiday + day_of_week + hour + cab_combo + surge_multiplier,
                      data = merged_data,
                      family = Gamma(link = "log"))
summary(glm_model_price)
appraise(glm_model_price)
PseudoR2(glm_model_price, which = NULL)
```

```{r}
#Validating GLM
set.seed(42)
train_control <- trainControl(method = "cv", number = 10)

glm_kfold <- train(price ~ holiday + day_of_week + hour + cab_combo + surge_multiplier,
                      data = merged_data,
                      method="glm",
                      family = Gamma(link = "log"),
                      trControl = train_control)
print(glm_kfold)
```

## Results

**Research Question 1:**

The GAM explained approximately 92 percent of the deviance in ride prices, with an adjusted R squared of approximately 0.90. The dominant predictors were distance, cab type, and service name rather than weather conditions.
 
**Figure #: Estimated Smooth Effects of Weather Variables on Log(Price):**
 
Figure # displays the estimated smooth effects for each of the six weather variables on the log price scale, with shaded areas representing 95 percent confidence intervals and rug marks showing the distribution of observed values. Among the weather smooth terms, only cloud cover reached statistical significance with a p value of 0.002. Temperature showed a non-linear pattern but did not reach significance at the 0.05 level with a p value of 0.11. Atmospheric pressure was close to significance with a p value of 0.09 but did not cross the threshold. Wind, humidity, and rain showed no significant effects after controlling for the structural covariates, with their confidence intervals overlapping zero throughout. The practical magnitude of even the significant cloud cover effect was small. The smooth curves for all weather variables varied by less than 0.05 on the log price scale, translating to price differences of roughly 5 percent or less across the observed range of each variable.
 
**Figure #: GAM Diagnostic Plots:**
 
Figure # presents the standard GAM diagnostic checks. The QQ plot of deviance residuals in the top left panel was approximately linear, indicating that the Gamma distribution assumption was reasonable. The residuals versus linear predictor plot in the top right panel showed no strong systematic patterns, suggesting that the model adequately captured the relationship between covariates and price. The histogram of residuals in the bottom left panel confirmed an approximately symmetric distribution centred near zero. The response versus fitted plot in the bottom right panel demonstrated that the model captured the general price structure well, with points clustering around the diagonal. Overall, the Gamma family handled the skewed distribution appropriately and the diagnostics did not reveal major model misspecification.
 
In summary, only cloud cover showed a statistically significant effect on ride prices among the six weather variables, but even this effect was practically negligible compared to structural factors such as distance, service type, and route. This suggests that during the observed period, the pricing algorithms of both Uber and Lyft were not meaningfully influenced by weather conditions.

**Research Question 2:**

The McFadden pseudo R squared for the logistic GLM was 0.045. While this may appear low, pseudo R squared values for logistic regression are not directly comparable to OLS R squared. McFadden (1979) noted that values between 0.2 and 0.4 represent excellent fit in discrete choice models. A value of 0.045 is common for models predicting rare events, as the baseline rate of surge occurrence was only 8.2 percent. The model's value lies in identifying which predictors are associated with surge rather than in predicting individual ride outcomes.
 
The analysis of deviance showed that pickup location was by far the strongest predictor of Lyft surge pricing, contributing a deviance reduction greater than all other predictors combined. Destination contributed a deviance reduction of approximately 32, and time period contributed approximately 12.5. Among the weather variables, only atmospheric pressure reached statistical significance with a deviance reduction of approximately 6, though even this effect was small in practical terms.
 
One notable finding involved service type. After excluding Shared rides, the service type variable contributed exactly zero deviance to the model. This means that Lyft applies the surge multiplier uniformly across all remaining service tiers. Whether a rider orders a regular Lyft, Lyft XL, Lux, or Lux Black, the same surge multiplier applies. Surge is therefore a platform level decision based on location and time rather than a service level decision.
 
**Figure #: Odds Ratios for Surge Occurrence (GLM):**
 
Figure # displays the odds ratios with 95 percent confidence intervals for the weather and time fixed effects from the GLM. Values greater than 1.0 indicate higher surge probability while the dashed line at 1.0 represents no effect. The odds ratios for weather variables all clustered tightly around 1.0, confirming negligible effects. Morning rush had an odds ratio of approximately 1.08 compared to midday, meaning surge odds were about 8 percent higher during morning commute hours. Evening rush and night showed similar but slightly smaller elevations. Rain had an odds ratio near 1.0 but with an extremely wide confidence interval because the vast majority of observations had zero rainfall. The model simply did not have enough rainy rides to estimate the effect of rain with any precision.
 
**Figure #: Binned Residual Plot (GLM):**
 
Figure # shows the binned residual plot for the logistic GLM, which is the standard diagnostic for binary response models. The plot divides the fitted probabilities into 40 bins and displays the average residual within each bin, with grey bands indicating approximate 95 percent confidence bounds. The majority of points fell within the expected bounds, indicating that the model fit was adequate overall. There were a few points near the higher fitted probability range that approached or slightly exceeded the bounds, suggesting the model may slightly underperform for rides with high predicted surge probability. However, the overall pattern did not indicate serious model misspecification.
 
For the GLMM robustness check, the random intercept for source had a standard deviation of 0.73 on the logit scale, confirming that substantial variation exists between pickup locations in terms of surge probability. This is consistent with the GLM finding that pickup location is the dominant predictor. The destination random intercept was near zero with a standard deviation of 0.04, indicating that the fixed effect dummy variables already fully capture destination level variation.
 
The critical result of the robustness check is the consistency of the fixed effect estimates across the two models. All weather and time coefficients differed by less than 0.001 on the logit scale between the GLM and GLMM. This confirms that regardless of whether location is treated as a fixed or random effect, the conclusions about weather and time are virtually identical. With only 12 unique locations per grouping variable and over 21,000 observations per group, fixed effects can reliably estimate each location's contribution without requiring the shrinkage that random effects provide (Bolker et al., 2009; Gelman and Hill, 2007).
 
These results did not support our hypothesis. Surge pricing is not primarily driven by temporal factors or weather. Instead, it is dominated by pickup location. The practical implication is that a rider's chances of encountering surge depend far more on where they ride than on when they ride or what the weather is doing.
**Research Question 3:**

**Data Summary Results GAM:**
The data summary of the GAM indicates that cab combo is the most significant covariate alongside surge multiplier. The EDF for the GAM is 2.5 which does indicate a degree of non-linearity, the baseline for linear being one. Our r-squared values also indicate that the GAM accounts for 81.6% of the deviance apparent in the model.

**Figure 1- Residuals Plots:**
The residual plots for our GAM give helpful insights into the model behaviour. The QQ Plot indicates that the model struggles at higher price values, this can be seen through the plot fitting the line well at the lower and middle values but curving upwards as the values increase. The residuals vs. Linear Predictor Plot and the Observed vs. Fitted Values Plot both have extreme banding. This we predict is due to how influential cab combo is on the price of Uber and Lyft ride prices. The bands on the Linear Predictor Plot do hover around zero however they peak in certain areas which indicates that the model is underestimating on certain high price rides. The observed vs. Fitted Values plot indicates there is variance being missed by the model. As described above the model accounts for 81.6% of the variance but there is still some that is not accounted for in this analysis. These gaps could be from covariates which were not included in the model for this research question. The Histogram of the Residuals is slightly right skewed as well which indicates that the mean is higher than the median, leading us to the conclusion we also drew from the QQ Plot which is that the model struggles predicting at higher price points.

**Data Summary Results GLM:**
Our GLM summary indicated that cab combo and surge multiplier are the most significant covariates for this model. The McFadden r-squared value we calculated is a 0.2469 which according to Stack Exchange applications of this function indicate a strong model fit (Reichenbach, 2014). 

**Figure 2- Residuals Plots:** 
The residual plots for our GLM are very similar to the residual plots of our GAM. Since there was not a significant difference in the model construction and there was only one continuous variable that was smoothed the plots are nearly identical. Which means the model for GLM also struggles as prices of rides increase as can be seen in the upwards curve of the QQ Plot. There are large variances in the Residuals vs. Linear Predictor Plot which indicate the model is underestimating the values of prices. The Observed vs. Fitted Values Plot indicates that there is quite a variance between the observed and predicted values. Since the GLM model also doesn’t account for all of the deviance it is likely that there are influences from other covariates in our dataset that are not taken into account for this question in particular. Finally, our histogram for the GLM is also positive right-skewed which indicates that the model is underfitting on the higher prices which is another confirmation for our observations above.

**K-Fold Cross Validation:**
To confirm this model we used k-fold cross validation with 10 folds. This was to further confirm the validity of the model by testing it against itself through withholding data. Withholding data for cross validation testing helps us maximize our dataset by training the model on some of our data and withholding the rest for validating the model performance on unseen data. The outcome of our k-fold cross validation was an RMSE of 3.98, r-squared of 81.75%, and an MAE of 2.95. The RMSE indicates that our model observed and fitted values were off on average by 3.98 dollars. Our r-squared score was 81.75%, which indicates how well the model performed on the validation set. And our MAE of 2.95 is the average difference between our price values (Farshad, 2024).

**Choosing GLM:**
We used both models to analyse the data prior to deciding on a singular model, but ultimately we chose the GLM. Prior to editing our analysis we had compared the GLM and GAM using AIC, which was not an accurate comparison since different models are not directly comparable. After reviewing the code and our analysis we looked at our existing code summaries and model covariates practically. There was only one continuous covariate in our model (hour), and although the EDF in the GAM was 2.5 indicating non-linearity hour is not significant enough to warrant keeping the GAM.  As discussed in the course it makes sense to choose the more simple model in situations like these when results and performance are similar. 

## Conclusion:

## Limitations:

## References
