---
title: "Project Report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Project Report}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

*Note:* Before running the code please generate the
merged_data_cleaned.csv. To do this ensure you have saved the raw data
files from the repository and then run Cleaning And Merging The Data.R
to generate the merged_data_cleaned.csv. Due to file size limitations we
were unable to upload the compiled merged_data_cleaned.csv file
directly.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data.table)
library(UberLyftAnalysis)
library(dplyr)
library(lubridate)
library(ggplot2)
library(gratia)
library(mgcv)
library(caret)
library(DescTools)

merged_data <- read.csv("../merged_data_cleaned.csv")
```

**Front Matter** 

Uber and Lyft Surge Pricing Analysis 
By Lingyun Jiang (B00850695) and Sydney Parsons (B01068132) 
Data Analysis STAT 4620/5620 Winter 25-26 
Link to GitHub:
<https://github.com/SydneyEParsons/Uber_Lyft_SurgePricing_Analyis>

## Abstract (150 words)

This is an analysis of Uber and Lyft Pricing in Boston Massachusetts in
November to December of 2018 (Munde, 2019). In it we cover three main
research questions with the goal of determining the overall impacts on
pricing for Uber and Lyft.

##Keywords 

Uber, Lyft, Pricing, Surge, Holiday, Rushhour, DOW, Weather, Distance


##Introduction Research questions and background information.

**Research Question 1:**

**Research Question 2:**

**Research Question 3:**



## Data Description:

-   Data points description here\*\*


**Variables:**



### EDA: Research Question 1

**Figure 1:**

```{r}

```

**Figure 2:**

```{r}

```

**Figure 3:**

```{r}

```

### EDA: Research Question 2

**Figure \#**

```{r}

```

### EDA: Research Question 3

```{r}

```

## Methods

**Research Question 1:** 
*Discuss methods chosen and why*

**Research Question 2:** 
*Discuss methods chosen and why*

**Research Question 3:** 
*Discuss methods chosen and why*


## Analysis

### Research Question #1 
(*Insert code and validation into the vignette*)

```{r}

```

### Research Question #2 
(*Insert code and validation into the vignette*)

```{r}

```

### Research Question #3 
(*Insert code and validation into the vignette*)

```{r}

```

## Results

**Research Question 1:**Discuss the analysis and the results why or why
not it worked\*

(*Insert code and validation for the analysis into the vignette*)

```{r}

```

**Research Question 2:**Discuss the analysis and the results why or why
not it worked\*

(*Insert code and validation for the analysis into the vignette*)

```{r}

```

**Research Question 3:**Discuss the analysis and the results why or why
not it worked\*

(*Insert code and validation for the analysis into the vignette*)

```{r}

```

## Conclusion:

## Limitations:

## References
