---
title: "Project Report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Project Report}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

**Front Matter**

Uber and Lyft Surge Pricing Analysis By Lingyun Jiang (B00850695) and
Sydney Parsons (B01068132) 

Data Analysis STAT 4620/5620 Winter 25-26

Link to GitHub:
<https://github.com/SydneyEParsons/Uber_Lyft_SurgePricing_Analyis>


*Note:* Before running the code please generate the
merged_data_cleaned.csv. To generate this file ensure you have saved the
raw data files from the repository here
(<https://github.com/SydneyEParsons/Uber_Lyft_SurgePricing_Analyis/tree/main>)
and then run Cleaning And Merging The Data.R found in the R folder here
on GitHub
(<https://github.com/SydneyEParsons/Uber_Lyft_SurgePricing_Analyis/blob/main/R/Cleaning%20And%20Merging%20The%20Data.R>).
Due to file size limitations we were unable to upload the compiled
merged_data_cleaned.csv file directly.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup}
library(dplyr)
library(lubridate)
library(ggplot2)
library(gratia)
library(mgcv)
library(caret)
library(DescTools)
library(scales)
library(tidyr)
library(lme4)
```

## Abstract

This is an analysis of Uber and Lyft Pricing in Boston Massachusetts for
23 days in late November to early December of 2018 (Munde, 2019). In
this paper we cover three main research questions with the goal of
determining the overall impacts on pricing for Uber and Lyft.

Ultimately we determined that the primary impacts on Uber and Lyft
pricing are their cab types and their ride distance. The pricing for
each of these ride shares are very segmented and stable within their
type boundaries. Distance is the variable factor that changes the most,
as riders choose rides that are further and shorter distances that will
impact their ride cost. Although their prices do change with influences
from our other variables mentioned, they are not significant enough when
compared to the influence of cab type and distance on price.

## Keywords

Uber, Lyft, Pricing, Surge, Holiday, Rushhour, DOW, Weather, Distance

## Introduction Research questions and background information.

After reviewing the Uber and Lyft pricing dataset by Munde on Kaggle we
chose to look more deeply at the impacts on Uber and Lyft ride pricing.
As users of both ride share apps ourselves we thought it would be an
interesting and practical analysis for an app people use daily. Our
analysis plans to break down the variables that are included in the
dataset and determine which of them most significantly impact price.

Ride-hailing platforms like Uber and Lyft have become a central part of
urban transportation. Their pricing, however, remains opaque to most
riders. Prices fluctuate based on factors that are not always visible,
and the platforms themselves reveal little about how their algorithms
work. This raises a practical question: what actually drives the price?

One common assumption is that weather plays a role. On a rainy evening
or a freezing morning, more people request rides while fewer drivers are
willing to work. Basic supply-and-demand logic suggests this should push
prices up. Yet whether this intuition holds in practice, and whether the
effect is large enough to matter, has not been thoroughly tested with
real data.

A related question concerns surge pricing. Lyft applies a visible surge
multiplier when demand exceeds supply in a given area. Understanding
what triggers this multiplier is useful for riders who want to avoid
paying more and for researchers studying platform pricing behavior.
Weather, time of day, day of week, and geographic location are all
plausible candidates.

We address three research questions. The first asks whether weather
conditions have a significant and possibly non-linear effect on ride
price, after controlling for distance, route, cab type, and service
level. The second asks what factors, including weather, time, and
location, are associated with the occurrence of Lyft surge pricing. The
third, conducted by the second author, examines the impact of rush hours
and holidays on pricing for both platforms.

The analysis combines a Generalized Additive Model (GAM) for the price
question with a logistic Generalized Linear Model (GLM) and a
Generalized Linear Mixed Model (GLMM) robustness check for the surge
question. Each model was chosen to match the structure of the response
variable and the nature of the research question, as described in the
relevant sections below.

**Research Question 1:** After controlling for distance, route, cab
type, and service level, do weather conditions (temperature, wind,
humidity, cloud cover, pressure, and rain) have a significant effect on
ride price, and is the relationship non-linear?

Our hypothesis was that adverse weather conditions, such as rain or
sub-freezing temperatures, would trigger price increases that do not
occur under mild conditions. We also expected this relationship to be
non-linear rather than linear. A drop from 50°F to 40°F might have
little effect, but a drop from 35°F to 25°F, where conditions become
genuinely uncomfortable, could produce a larger price shift. If such
threshold effects exist, a model that allows flexible curve shapes would
be necessary to detect them.

**Research Question 2:** For Lyft rides, what factors, including weather
conditions, time of day, and route, are associated with the occurrence
of surge pricing?

We hypothesized that surge pricing would be primarily driven by temporal
factors such as rush hours and by weather conditions such as rain or low
temperature, rather than by geographic location alone. The reasoning was
that adverse weather reduces driver supply relative to rider demand,
which should trigger the platform’s surge algorithm. We also expected
that certain time periods, particularly morning and evening commutes,
would show elevated surge rates.

**Research Question 3:** What is the impact of holiday, rush hour, day
of week, and cab type on the prices for Uber and Lyft rides during the
specified time period?

For the hypothesis for our final research question we predict that there
will be an impact on pricing related to rush hours, holidays and days of
the week. We assume this because time of day and day of the week would
be different depending on rush hours, normal hours, and if people are
commuting to work. This thinking directly relates to our assumptions on
holidays as well that if there are holidays people will be travelling or
participating in holiday events. Leading us to expect to see increases
in pricing during these periods.

## Data Description:

```{r}

# Load merged data
merged_data <- read.csv("../merged_data_cleaned.csv")
merged_df <-merged_data
##Research Question #1

# Replace rain NA with 0 (NA indicates no rain was recorded)
merged_df$rain[is.na(merged_df$rain)] <- 0

# Remove rows with missing price
merged_df <- merged_df %>% filter(!is.na(price))

# Create useful derived variables
# Data was collected in Boston (EST = UTC-5 in Nov-Dec 2018)
merged_df <- merged_df %>%
  mutate(
    time_stamp  = as.POSIXct(time_stamp, tz = "UTC"),
    hour_ts     = as.POSIXct(hour_ts, tz = "UTC"),
    time_local  = with_tz(time_stamp, tzone = "America/New_York"),
    hour_of_day = hour(time_local),
    day_of_week = lubridate::wday(time_local, label = TRUE),
    cab_label   = paste(cab_type, name, sep = "-"),
    route       = paste(source, "->", destination),
    surge_binary = ifelse(surge_multiplier > 1, 1, 0)
  )

cat("Dataset dimensions:", nrow(merged_df), "rows x", ncol(merged_df), "columns\n")
cat("Date range:", as.character(min(merged_df$time_stamp)),
    "to", as.character(max(merged_df$time_stamp)), "\n")


##Research Question #2
# Handle missing values
merged_df$rain[is.na(merged_df$rain)] <- 0
merged_df <- merged_df %>% filter(!is.na(price))

# Create derived variables
# Convert to Boston local time (EST = UTC-5 in Nov-Dec 2018) before extracting hour/day
merged_df <- merged_df %>%
  mutate(
    time_stamp   = as.POSIXct(time_stamp, tz = "UTC"),
    time_local   = with_tz(time_stamp, tzone = "America/New_York"),
    hour_of_day  = hour(time_local),
    # Bin hours into meaningful time periods to capture non-linear hourly patterns
    time_period  = factor(case_when(
      hour_of_day >= 6  & hour_of_day < 10 ~ "morning_rush",
      hour_of_day >= 10 & hour_of_day < 16 ~ "midday",
      hour_of_day >= 16 & hour_of_day < 20 ~ "evening_rush",
      TRUE                                  ~ "night"
    ), levels = c("midday", "morning_rush", "evening_rush", "night")),
    # Use unordered factor so R applies treatment (dummy) coding, not polynomial contrasts
    day_of_week  = factor(lubridate::wday(time_local, label = TRUE), ordered = FALSE),
    route        = paste(source, "->", destination),
    surge_binary = ifelse(surge_multiplier > 1, 1, 0)
  )


##Research Question #3
merged_data <- merged_data %>%
  mutate(
    hour_ts_boston = with_tz(ymd_hms(hour_ts, tz = "UTC"), tzone = "America/New_York"),
    date = as.Date(hour_ts_boston),
    hour = hour(hour_ts_boston),
    day_of_the_week = factor(wday(date, label = TRUE), ordered = FALSE),
    is_weekend = ifelse(wday(date) %in% c(1, 7), 1,0),
    rush_hour = case_when(
      is.na(hour) ~ NA_character_, #Some data points don't have a time
      hour >= 6 & hour <= 10 ~ "Morning Rush",
      hour >= 16 & hour <= 20 ~ "Evening Rush",
      TRUE ~ "Normal Hours"
    ),
    holiday = case_when(
      date == as.Date("2018-11-26") ~ "Cyber Monday",
      date >= as.Date("2018-12-02") & date <= as.Date("2018-12-10") ~ "Hanukkah",
      TRUE ~ "None"
    ),
    cab_combo = paste(cab_type, name, sep = "-")
  )

merged_data <- merged_data %>% filter(!is.na(hour))

lyft_data <- merged_data %>%
  filter(cab_type == "Lyft")

raw <- read.csv("../merged_data_cleaned.csv") 
#Shows how many data points don't have a time
sum(!grepl(":", raw$hour_ts_boston))

class(merged_data$hour_ts_boston)
head(merged_data$hour_ts_boston)
```

**Variables:**

-   **Weather:** Temperature (°F), Clouds (0-1, fraction of sky
    covered), Humidity (Percentage), Pressure (Millibars), Rain
    (Inches), Wind (mph).
-   **Distance:** Ride distance is measured in Miles.
-   **Pickup Location:** Includes: Black Bay, Northeastern University,
    Fenway, Boston University, Theatre District, Beacon Hill, South
    Station, Financial District, West End, North Station, Haymarket
    Square, and North End.
-   **Holidays:** Hanukkah (December 2-10, 2018) and Cyber Monday
    (November 26, 2018).
-   **Rush hour:** Morning (6am - 10am), Midday (10 am - 4pm), Evening
    (4pm - 8pm), and Night (8pm - 6am).
-   **Surge:** 1.0 - 3.0, Uber remains at a constant surge multiplier of
    1 because there was no observed surge pricing for Uber in the
    construction of this dataset.
-   **Price:** Ride prices are measured in US Dollars (USD).
-   **Uber:** One of the ride hailing apps being analysed in this
    dataset.
-   **Lyft:** The other ride hailing app is being analysed in this
    dataset.
-   **Cab type:** Includes: Lux, Lux Black, Lux Black XL, Lyft, Lyft XL,
    Shared, Black, Black SUV, UberPool, UberX, UberXL, WAV.
-   **Time:** Time was displayed as Boston Time zone instead of the
    original Unix so it can show the precise time calculations for the
    area.
-   **Day of the week:** Days of the week: Monday, Tuesday, Wednesday,
    Thursday, Friday, Saturday, Sunday.
-   **Cab_combo:** This combines Uber and Lyft with their respective cab
    types so the data can be categorized by ride hailing app and cab
    type all in one. Lyft-Lux, Lyft-Lux Black, Lyft-Lux Black XL,
    Lyft-Lyft, Lyft-Lyft XL, Lyft-Shared, Uber-Black, Uber-Black SUV,
    Uber-UberPool, Uber-UberX, Uber-UberXL, Uber-WAV

This report uses a dataset of approximately 637,000 ride-hailing trips
collected in the Boston area over a 23-day window from late November to
mid-December 2018 (Munde, 2019). The data, obtained from Kaggle,
includes ride-level information from both Uber and Lyft alongside hourly
weather observations for six variables: temperature, wind speed,
humidity, cloud cover, atmospheric pressure, and rainfall. As noted in
our proposal we read a paper published on this dataset in 2023 by Huang
which detailed the units used in the measurements (Huang, 2023). In
addition to reading the paper we sought out confirmations of the units
on our own by reviewing how different weather patterns are measured,
conditions in Massachusetts and standard measurement units in the United
States.

To clean the data we started by checking the number of unique values for
timestamps and for duplicates of the data. No duplicates were found but
there were much less time stamps than rows in our dataset. This led us
to converting cab ride times from Unix to POSIXct so we could have them
categorized to their nearest hour. Since the dataset focused on Uber and
Lyft we removed the values for Taxi that were included in the data. The
number of data points for Taxi was low compared to Uber and Lyft
datapoints and since Taxis are not a ride hailing app as specified by
our research questions we ultimately removed them during cleaning. To
finish off our data cleaning we merged the cab rides dataset with the
weather values to create our clean merged data file.

Throughout our project we used many of the tools from our statistical
analysis toolbox. After identifying our research questions and reviewing
the data we conducted our exploratory analyses. This process included
many different visualizations and plots of the data alone and compared
to other values. Our exploratory analysis was split into each of our
individual research questions; this helped us focus specifically on the
covariates important to that specific question.

### EDA: Research Question 1

**Figure 1: Distribution of Ride Prices**

```{r fig.width=8, fig.height=8}
ggplot(merged_df, aes(x = price)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Ride Prices",
       x = "Price (USD)", y = "Count") +
  theme_minimal()
```

For our first research question we began by examining the distribution
of ride prices. As shown in Figure 1, the histogram revealed a strong
right skew with the majority of rides concentrated at lower prices and a
long tail extending toward more expensive service tiers. This pattern
suggested that a model assuming normally distributed errors would not be
appropriate and that the Gamma distribution would better capture the
shape of the response variable.

**Figure 2: Temperature vs. Price (Log Scale) by Platform:**

```{r fig.width=8, fig.height=8}
set.seed(42)
sample_df <- merged_df %>% slice_sample(n = 20000)

ggplot(merged_df, aes(x = temp, y = price)) +
  geom_point(alpha = 0.02, size = 0.3, color = "grey40") +
  geom_smooth(data = sample_df, method = "gam", formula = y ~ s(x),
              se = TRUE, color = "steelblue", linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~ cab_type) +
  labs(title = "Temperature vs. Price (Log Scale) by Platform",
       x = "Temperature (F)", y = "Price (USD, log scale)") +
  theme_minimal()
```

We then plotted temperature against price on a log scale with a GAM
smoother overlaid on a random subsample. As shown in Figure 2, the
smoother revealed a slight non-linear trend, but the effect was visually
minor compared to the spread of the data. This indicated that while a
non-linear relationship may exist, its practical magnitude is small
relative to other sources of price variation. The horizontal banding
pattern visible in both panels reflects the distinct pricing tiers of
different service levels (e.g., UberPool, UberX, Black), which produce
discrete price clusters. Within each tier, prices show little variation
across the temperature range.

**Figure 3: Correlation Between Weather and Price by Service Type:**

```{r fig.width=8, fig.height=8}
cor_results <- merged_df %>%
  group_by(cab_label) %>%
  summarise(
    Temperature = cor(price, temp,     use = "complete.obs"),
    Clouds      = cor(price, clouds,   use = "complete.obs"),
    Pressure    = cor(price, pressure, use = "complete.obs"),
    Humidity    = cor(price, humidity,  use = "complete.obs"),
    Wind        = cor(price, wind,     use = "complete.obs"),
    Rain        = cor(price, rain,     use = "complete.obs"),
    .groups = "drop"
  )

cor_long <- cor_results %>%
  pivot_longer(-cab_label, names_to = "Weather Variable", values_to = "Correlation")

ggplot(cor_long, aes(x = `Weather Variable`, y = cab_label, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 3)), size = 2.8) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  labs(title = "Correlation Between Weather and Price by Service Type",
       subtitle = "Unadjusted Pearson correlations (not controlling for distance or route)",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Figure 3 presents the unadjusted Pearson correlations between weather
variables and price for each service type. The correlations were
uniformly weak across all service types with most values falling between
negative 0.01 and positive 0.01. Temperature showed the strongest
individual correlation, but it remained small in absolute terms. These
patterns indicated that weather variables are unlikely to be dominant
price drivers, motivating our use of a model that can detect subtle and
potentially non-linear effects.

### EDA: Research Question 2

```{r fig.width=8, fig.height=8}
# Confirm Uber surge is always 1, only Lyft has variation
merged_df %>%
  group_by(cab_type) %>%
  summarise(
    n_rides      = n(),
    unique_surge = n_distinct(surge_multiplier),
    min_surge    = min(surge_multiplier),
    max_surge    = max(surge_multiplier),
    pct_surge    = round(mean(surge_multiplier > 1) * 100, 2),
    .groups = "drop"
  )
```

For our second research question we focused on the Lyft subset of the
data because Uber showed a constant surge multiplier of 1.0 throughout
the entire dataset. Shared rides were excluded as they also showed no
surge variation, which would cause estimation problems in a logistic
regression.

```{r fig.width=8, fig.height=8}
# Focus on Lyft only, exclude Shared rides
lyft_df <- merged_df %>%
  filter(cab_type == "Lyft", name != "Shared")

cat("Lyft rides (excl. Shared):", nrow(lyft_df), "\n")
cat("Surge rate:", round(mean(lyft_df$surge_binary) * 100, 2), "%\n\n")
cat("Surge multiplier distribution:\n")
table(lyft_df$surge_multiplier)
```

**Figure 4: Surge Rate by Pickup Location:**

```{r fig.width=8, fig.height=8}
source_surge <- lyft_df %>%
  group_by(source) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(source_surge, aes(x = reorder(source, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Pickup Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

As shown in Figure 4, there was substantial variation in surge rates by
pickup location, ranging from 1.8% at North End to 13.4% at Back Bay.
This suggested that pickup location factors play a significant role in
surge occurrence.

**Figure 5: Surge Rate by Dropoff Location:**

```{r fig.width=8, fig.height=8}
dest_surge <- lyft_df %>%
  group_by(destination) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(dest_surge, aes(x = reorder(destination, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Dropoff Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

Figure 5 presents a similar analysis for dropoff locations. The
variation across destinations was also present though somewhat less
pronounced than for pickup locations, indicating that where a ride
begins may matter more than where it ends for surge pricing.

**Figure 6: Hourly Surge Rate (Lyft):**

```{r fig.width=8, fig.height=8}
hour_surge <- lyft_df %>%
  group_by(hour_of_day) %>%
  summarise(surge_rate = mean(surge_binary), n_rides = n(), .groups = "drop")

ggplot(hour_surge, aes(x = hour_of_day, y = surge_rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Hourly Surge Rate (Lyft)",
       x = "Hour of Day", y = "Surge Rate") +
  theme_minimal()
```

Figure 6 showed that surge rates peaked around 8am at approximately
9.5%, with a secondary peak around 3pm. The lowest surge rate occurred
around 11am at approximately 7.0%. The overall range was narrow,
spanning only about 2.5 percentage points, indicating that time of day
has a limited effect on surge occurrence.

**Figure 7: Weather Conditions: Surge vs. Non-Surge:**

```{r fig.width=8, fig.height=8}
weather_vars <- c("temp", "wind", "humidity", "clouds", "pressure", "rain")

# Standardize weather variables for comparable scale
lyft_std <- lyft_df %>%
  mutate(across(all_of(weather_vars), ~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE))) %>%
  mutate(Surge = ifelse(surge_binary == 1, "Surge", "No Surge")) %>%
  select(Surge, all_of(weather_vars)) %>%
  pivot_longer(-Surge, names_to = "Variable", values_to = "Value")

lyft_means <- lyft_std %>%
  group_by(Surge, Variable) %>%
  summarise(mean_val = mean(Value, na.rm = TRUE), .groups = "drop")

ggplot(lyft_means, aes(x = Variable, y = mean_val, fill = Surge)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7),
           width = 0.6, alpha = 0.8) +
  scale_fill_manual(values = c("No Surge" = "grey60", "Surge" = "steelblue")) +
  labs(title = "Weather Conditions: Surge vs. Non-Surge (Lyft)",
       subtitle = "Standardized means (z-scores); small differences suggest weather is not the 
       primary surge driver",
       x = NULL, y = "Standardized Mean") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

Figure 7 compared standardized weather variable means between surge and
non-surge rides. The differences were extremely small across all weather
variables, providing early evidence that weather is not the primary
trigger for surge pricing.

**Figure 8: Surge Rate by Time Period and Day of Week:**

```{r fig.width=8, fig.height=8}
time_day_surge <- lyft_df %>%
  group_by(time_period, day_of_week) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  )

ggplot(time_day_surge, aes(x = day_of_week, y = time_period, fill = surge_rate)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = percent) +
  labs(title = "Surge Rate by Time Period and Day of Week (Lyft)",
       x = NULL, y = NULL, fill = "Surge Rate") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Figure 8 presents a heatmap of surge rates across time periods and days
of the week. The heatmap confirmed that temporal patterns exist but are
secondary to location effects. Certain combinations of time period and
day showed slightly elevated surge rates, but no single temporal factor
dominated the way that pickup location did.

### EDA: Research Question 3

**Figure 9: Price Distribution By Cab Type:**

```{r fig.width=8, fig.height=8}
ggplot(merged_data, aes(x = cab_combo, y = price, fill = cab_combo)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~ cab_type, scales = "free_x") +
  scale_y_continuous(limits = c(0, 60)) +
  labs(title = "Price Distribution by Cab Type",
       x = "Cab Type", y = "Price") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

To begin our EDA plots for our third research question we plotted our
cab combo data compared to price. This was a starting point as we
further breakdown the pricing analysis in the coming EDA plots. In this
EDA plot we used boxplots to demonstrate this distribution and median of
the pricing data per each category. As you can see from the
visualization the plots show very segmented data. Both ride hailing apps
have cab-types that are very regimented with their pricing, there is
limited overlap between the pricing of different cab types.

**Figure 10: Analyze Hourly Price on Normal Days of the Week:**

```{r fig.width=8, fig.height=8}
#Analyze hourly price on normal days of the week
normal_rush_dayofweek_summary <- merged_data %>%
    filter(holiday=="None") %>%
    group_by(hour, day_of_the_week, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )

ggplot(data=normal_rush_dayofweek_summary %>% filter(cab_type == "Lyft"), 
       mapping =aes(x=hour, y=mean_price, color=day_of_the_week)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Analyze hourly price on normal days of the week
", x="hour", y="mean_price") 

ggplot(data=normal_rush_dayofweek_summary %>% filter(cab_type == "Uber"), 
       mapping =aes(x=hour, y=mean_price, color=day_of_the_week)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Analyze Hourly Price on Normal Days of the Week
", x="Hour", y="Mean Price") 
```

These line graphs show the mean pricing for the different cab combos
during the days of the week. The lines overall remain very stable but
there is greater variation with pricing in the higher price hailing
tiers, specifically in Lyft-Lux Black, Lyft-Lux Black XL, Uber-Black,
and Uber-Black SUV. These variances are not significant but it is clear
there is more noise in these combos than the others.

**Figure 11: Compare Day of the Week, Price and Cab Combo:**

```{r fig.width=8, fig.height=8}
#Compare day of week, price and cab_combo
day_of_the_week <- merged_data %>%
  group_by(day_of_the_week, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(day_of_the_week, n = Inf)

ggplot(data=day_of_the_week, mapping =aes(x=day_of_the_week, y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)), size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Compare day of week, price and cab_combo
", x="Day of Week", y="Mean Price") +
  coord_flip() 

ggplot(data=day_of_the_week %>% filter(cab_type == "Lyft"), 
    mapping = aes(x=day_of_the_week, y=cab_combo, fill=mean_price)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2) +
labs(title = "Lyft - Compare day of week, price and cab_combo
", x="Day of Week", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data=day_of_the_week %>% filter(cab_type == "Uber"), 
    mapping = aes(x=day_of_the_week, y=cab_combo, fill=mean_price)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2) +
labs(title = "Uber - Compare day of week, price and cab_combo
", x="Day of Week", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Here we are focusing on the influence of the days of the week on the
pricing of ride hailing platforms Uber and Lyft. Our following EDA plots
is a colour coded bar plot and heat maps. These plots help illustrate
the relationship between the days of the week and our cab combo. As
apparent through the plots the pricing remains very consistent across
each day of the week and cab types.

**Figure 12: Price Distribution by Day of The Week:**

```{r fig.width=8, fig.height=8}
ggplot(merged_data, aes(x = day_of_the_week, y = price, fill = day_of_the_week)) +
  geom_violin(trim = FALSE) +  
  geom_boxplot(width = 0.1, outlier.shape = NA) +  
  facet_wrap(~ cab_type) +
  scale_y_continuous(limits = c(0, 60)) +  
  labs(title = "Price Distribution by Day of The Week",
       x = "Day of Week", y = "Price") +
  theme_minimal() +
  theme(legend.position = "none") 
```

Since more variation was expected in the data as highlighted in our
hypothesis, we plotted an additional EDA plot as a violin plot. This
plot helps show the quantity of rides at each price point through the
density of the waves. In our presentation it was highlighted that they
were very uniform and that we should investigate further. After
analysing the code we realized the smoothing we applied to the pots was
distorting the final display. After removing that we were able to see
the density of the prices more clearly. However, this plot reaffirms the
lack of variation in pricing throughout the days of the week as seen in
our initial EDA plots above.

**Figure 13: Compare Weekday and Weekends to Price and Cab Combo:**

```{r fig.width=8, fig.height=8}
#Compare weekday and weekends to price and cab_combo
weekend_summary <- merged_data %>%
  group_by(is_weekend, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(weekend_summary)

ggplot(data=weekend_summary, 
       mapping =aes(x=factor(is_weekend, labels=c("Weekday", "Weekend")), 
                    y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)),  size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Compare weekday and weekends to price and cab_combo", 
     x="Is Weekend", y="Mean Price") +
  coord_flip() 
```

With little variation in the days of the week we analysed the data
further. By plotting the week day against the weekend we aimed to try to
identify price variations on a more general basis but this plot also did
not show any notable variation in pricing from weekday to weekend.

**Figure 14: Price Distribution By Holiday:**

```{r fig.width=8, fig.height=8}
#Comparing weekend and weekday pricing to holiday pricing
holiday_summary_detailed <- merged_data %>%
  group_by(holiday, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(holiday_summary_detailed)

ggplot(merged_data, aes(x = holiday, y = price, fill = holiday)) +
  geom_violin(trim = FALSE) +  
  geom_boxplot(width = 0.1, outlier.shape = NA) +  
  facet_wrap(~ cab_type) +
  scale_y_continuous(limits = c(0, 60)) +  
  labs(title = "Price Distribution by Holiday",
       x = "Holiday", y = "Price") +
  theme_minimal() +
  theme(legend.position = "none") 

ggplot(data=holiday_summary_detailed, 
       mapping =aes(x=factor(holiday, labels=c("Cyber Monday", "Hanukkah", "None")), 
                    y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)),  size = 2.5, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Comparing Holiday Pricing to Cab Combo", x="Holiday", y="Mean Price") +
  coord_flip() 
```

For price distribution by holiday, we presumed traffic would be higher
on these days due to people travelling for the holiday and having time
off. However, as seen in the violin plots and the bar plots above that
was not the case. This may indicate that the pricing was changing
slightly between cab-type segments on Lyft rides. However, Uber does
remain very constant in its pricing regardless of holiday. The bar graph
also shows the Lyft ride prices being more incremental than Uber ride
pricing which remains very stable.

**Figure 15: Comparing Cost of Rides on Holidays:**

```{r fig.width=8, fig.height=8}
#Compare time of day price on normal days to time of day on holiday
normal_time_holiday_time_summary <- merged_data %>%
    filter(is_weekend==0) %>%
    group_by(hour, holiday, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )

ggplot(data=normal_time_holiday_time_summary %>% filter(cab_type == "Lyft"), 
       mapping =aes(x=hour, y=mean_price, color=holiday)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Compare Cost of Rides on Holidays", x="hour", y="mean_price") 

ggplot(data=normal_time_holiday_time_summary %>% filter(cab_type == "Uber"), mapping =aes(x=hour, y=mean_price, color=holiday)) +
geom_line() +
facet_wrap(~ cab_combo) +
labs(title = "Compare Cost of Rides on Holidays", x="hour", y="mean_price")
```

To further visualise the pricing of Uber and Lyft on holidays we plotted
them on line graphs. This provides an additional visualization outside
of the violin plots. It is clear from these plots that the prices remain
stable with small peaks later in the days presumably around the evening
rush hour period.

**Figure 16: Comparing Rush Hour Price for Weekdays to Price on Normal Weekends:**

```{r fig.width=8, fig.height=8}
#Comparing rush hour price for weekdays to price on weekends
rush_weekdays_normal_weekends_summary <- merged_data %>%
  group_by(rush_hour, is_weekend, cab_combo, cab_type) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )
print(rush_weekdays_normal_weekends_summary)

ggplot(data=rush_weekdays_normal_weekends_summary, 
       mapping =aes(x=rush_hour, y=mean_price, fill=cab_combo)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_price, 2)),  size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ is_weekend,
  labeller = labeller(is_weekend = c("0" = "Weekday", "1" = "Weekend"))) +
labs(title = "Comparing rush hour price for weekdays to price on normal weekends
", x="Rush Hour", y="Mean Price") +
  coord_flip() 

ggplot(data=rush_weekdays_normal_weekends_summary %>% filter(cab_type == "Lyft"),
    mapping = aes(x=rush_hour, y=cab_combo, fill=mean_price)) +
geom_tile() +
facet_wrap(~ is_weekend, labeller = labeller(is_weekend = c("0" = "Weekday", "1" = "Weekend"))) +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2.5) +
labs(title = "Lyft - Comparing rush hour price for weekdays to price on normal weekends
", x="Rush Hour", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data=rush_weekdays_normal_weekends_summary %>% filter(cab_type == "Uber"),
    mapping = aes(x=rush_hour, y=cab_combo, fill=mean_price)) +
geom_tile() +
facet_wrap(~ is_weekend, labeller = labeller(is_weekend = c("0" = "Weekday", "1" = "Weekend"))) +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_price, 2)),  size = 2.5) +
labs(title = "Uber - Comparing rush hour price for weekdays to price on normal weekends
", x="Rush Hour", y="Mean Price") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

In this EDA section we plotted the comparison of rush hour to weekdays
and weekends on a bar plot and a heatmap categorized by cab-type. These
plots demonstrates that the prices for rides remain steady across normal
hours, morning rush and evening rush on both evenings and weekends. The
consistency in the bar graph values and teh heatmap values further
confirm the limited change in pricing across the varying rush hours on
weekdays and weekends.

**Figure 17: Price Distribution by Rush Hour:**

```{r fig.width=8, fig.height=8}
ggplot(merged_data, aes(x = rush_hour, y = price, fill = rush_hour)) +
  geom_violin(trim = FALSE) +  
  geom_boxplot(width = 0.1, outlier.shape = NA) +  
  facet_wrap(~ cab_type) +
  scale_y_continuous(limits = c(0, 60)) +  
  labs(title = "Price Distribution by Rush Hour",
       x = "Rush Hour", y = "Price") +
  theme_minimal() +
  theme(legend.position = "none") 
```

The following violin plot demonstrates the price distribution of Uber
and Lyft apps during our three categorized rush hour periods. As can be
seen in the plot the price remains very similar across all three rush
hour periods. The wider sections of the violin plots in this EDA
indicate that there are more rides at that price point. Both Uber and
Lyft have very specific shapes that do not change as rush hour changes.

**Figure 18: Compare Cab Combo, Holiday and Uber & Lyft Surge
Multiplier:**

```{r fig.width=8, fig.height=8}
#Compare cab_combo, holiday and surge multiplier
holiday_surge_summary <- merged_data %>%
  group_by(holiday, cab_type, cab_combo) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE),
    mean_surge = mean(surge_multiplier, na.rm = TRUE),
    median_surge = median(surge_multiplier, na.rm = TRUE)
  )

ggplot(data=holiday_surge_summary, mapping =aes(x=cab_combo, y=mean_surge, fill=holiday)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
geom_text(aes(label=round(mean_surge, 2)),  size = 2, position=position_dodge(width=0.9)) +
facet_wrap(~ cab_type) +
labs(title = "Compare Cab Combo, Holiday & Surge Multiplier", x="Cab Type", y="Surge Pricing") +
  coord_flip() 
```

Surge multiplier was highlighted in our earlier research questions but
we did also briefly plot it here to test its influence in our third
research question. Uber remains constant because there was no observed
surge price in this collected dataset. You can see the surge prices for
Uber and Lyft here but a bar plot was not the most ideal for comparing
surge across cab combo and holiday which is why we pivoted to a heatmap
as can be seen in the following plot.

**Figure 19: Lyft and Uber Compare Cab Combo, Holiday and Surge
Multiplier:**

```{r fig.width=8, fig.height=8}
ggplot(data=holiday_surge_summary %>% filter(cab_type == "Lyft"), 
    mapping = aes(x=holiday, y=cab_combo, fill=mean_surge)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_surge, 2))) +
labs(title = "Lyft - Compare Cab Combo, Holiday and Surge Multiplier
", x="Holiday", y="Cab Combo") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(data=holiday_surge_summary %>% filter(cab_type == "Uber"), 
    mapping = aes(x=holiday, y=cab_combo, fill=mean_surge)) +
geom_tile() +
  scale_fill_gradient(low="white", high="blue") +
geom_text(aes(label=round(mean_surge, 2))) +
labs(title = "Uber - Compare Cab Combo, Holiday and Surge Multiplier
", x="Holiday", y="Cab Combo") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Here you can see a more clear breakdown of our surge visualization. A
heatmap for this data is more appropriate because it shows the
distribution through the colours and numbers on the plot in a much
cleaner presentation. As we can see from the values of the surge
multiplier for Lyft surge doesn’t change significantly over holidays
compared to non-holidays.

**Figure 20: Compare Pricing on Weekday Holidays to Rush Hour Pricing:**

```{r fig.width=8, fig.height=8}
#Compare pricing on weekday holidays to rush hour pricing
weekday_holiday_rush_summary <- merged_data %>%
    filter(is_weekend==0) %>%
    group_by(rush_hour, holiday, cab_type) %>%
  summarise(
    n = n(),
    mean_price = mean(price, na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  )

ggplot(data=weekday_holiday_rush_summary, mapping =aes(x=rush_hour, y=mean_price, fill=holiday)) +
geom_bar(
  stat = "identity",
  position = "dodge") +
facet_wrap(~ cab_type) +
geom_text(aes(label=round(mean_price, 2)),  size = 2, position=position_dodge(width=0.9)) +
labs(title = "Compare Pricing on Weekday Holidays to Rush Hour Pricing") +
  coord_flip()

ggplot(data=weekday_holiday_rush_summary, mapping =aes(x=rush_hour, y=holiday, fill=mean_price)) +
geom_tile() +
facet_wrap(~ cab_type) +
  scale_fill_gradient(low="white", high="red") +
geom_text(aes(label=round(mean_price, 2))) +
labs(title = "Compare pricing on weekday holidays to rush hour pricing
", x="Rush Hour", y="Holiday") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

For our comparison of weekday holidays to rush hour we initially plotted
a bar graph and then pivoted to a heatmap to better visualize the
values. The prices over all Uber and Lyft rides during all three rush
hour periods on holidays were very stable. There was a slight dip on the
price of Lyft rides in the evenings of Hannukah, and a slight increase
in price of Uber rides in the normal hours of Cyber Monday.

After completing our exploratory analyses we proceeded to create the
models for each of our research questions. The information provided from
our analysis helped us choose methods in an informed manner.

## Methods

**Research Question 1:** To assess the effect of weather on ride price
we used a Generalized Additive Model with a Gamma family and log link
function. This model was chosen based on three characteristics of the
response variable.

First, ride prices are strictly positive, which the Gamma distribution
naturally accommodates. Second, the distribution of prices is right
skewed with a concentration of rides at lower prices and a long tail of
expensive rides from premium service tiers. The Gamma distribution
handles this asymmetry well. Third, more expensive rides exhibit greater
price variability, which is consistent with the Gamma assumption that
variance is proportional to the squared mean.

The log link ensures that predicted prices remain positive and that
covariate effects act multiplicatively on the price scale. This is a
natural interpretation for pricing data because a one unit change in
temperature, for example, would scale the price by a constant percentage
rather than adding a fixed dollar amount.

We used smooth functions for the six weather variables to allow for
potentially non-linear relationships, while distance, cab type, service
name, source, and destination entered as parametric terms. The basis
dimensions were set to reflect the expected complexity of each
relationship. Temperature was assigned k equals 8 because it had the
widest range and the most potential for complex non-linearity. Clouds
and pressure were assigned k equals 5 for moderate flexibility. Wind,
humidity, and rain were each assigned k equals 4 because the data showed
limited variability in these variables or, in the case of rain, very
sparse observations. The effective degrees of freedom reported in the
model output indicate how much of this flexibility the data actually
require.

The model was fit using the bam function from the mgcv package (Wood,
2017), which is a computationally efficient version of gam designed for
large datasets. We used fast REML for smoothing parameter estimation
with the discrete option enabled to further accelerate computation. We
also specified select equals TRUE, which adds an extra shrinkage penalty
to each smooth term. This allows the model to shrink smooth terms to
zero effective degrees of freedom if the data do not support their
inclusion, providing a form of automatic variable selection for the
weather covariates.

The key assumptions of this model are that conditional on covariates,
prices follow a Gamma distribution with a constant shape parameter, and
that the smooth terms can adequately approximate the true underlying
relationships. We also assume that observations are conditionally
independent given the covariates. We acknowledge that repeated sampling
of the same routes over time may introduce correlation, which is
discussed further in the Limitations section.

**Research Question 2:** To model factors associated with surge pricing
we used a logistic regression with a binomial family and logit link,
treating surge occurrence as a binary outcome. A ride was classified as
having surge if the surge multiplier exceeded 1.0 and as not having
surge otherwise. This binary approach was chosen because the surge
multiplier in the Lyft data takes only a few discrete values, with the
vast majority of observations at 1.0 or 1.25. The extreme rarity of
higher surge levels made a continuous or ordinal model impractical.

The model includes weather variables consisting of temperature, wind,
humidity, cloud cover, pressure, and rain, as well as temporal variables
including time period and day of week, pickup location, dropoff
location, and service type as fixed effects. The time period was
categorised into four bins: morning rush from 6am to 10am, midday from
10am to 4pm, evening rush from 4pm to 8pm, and night from 8pm to 6am.

As a robustness check we also fit a Generalized Linear Mixed Model with
random intercepts for source and destination locations using the lme4
package (Bates et al., 2015). The purpose of this GLMM was to assess
whether treating locations as random effects, which partially pools
location level estimates and accounts for potential within location
correlation, would change our conclusions about the weather and time
predictors.

It is important to note that we do not use AIC or likelihood ratio tests
to compare the GLM and GLMM directly. These two models use different
likelihood formulations. The GLM maximises a conditional likelihood
while the GLMM with Laplace approximation uses an approximate marginal
likelihood that integrates over random effects. As discussed in the
course materials, standard AIC is only valid for comparing GLMM variants
when the goal is to account for correlation structures. Comparing across
model classes requires conditional AIC, which is beyond the scope of
this analysis. Instead, we evaluate the necessity of random effects by
examining the estimated random effect variances, the consistency of
fixed effect estimates between models, and the practical context of our
grouping structure.

The assumptions of the logistic GLM include independent observations
conditional on covariates and a linear relationship between covariates
and the log odds of surge. The GLMM additionally assumes that random
effects are normally distributed. Independence is approximate given the
repeated measures nature of the data, but with large sample sizes per
group this is unlikely to affect the direction of our conclusions.

**Research Question 3:** For research question three where we analysed
different effects on cab prices we chose to use a GLM. Initially we had
decided to compare a GLM and a GAM to see which one would perform best
on the dataset.

For the GAM, price was our response variable and the covariates used
were: holiday, day of the week, hour, cab combo and surge multiplier.
For our research question these were the covariates from our data that
represented our model the best. For the covariate of hour we applied a
smoothing function since it is continuous. The model distribution we
used was Gamma with a log link. This was chosen to best fit the data of
the response variable because it is positive and continuous, and the
link function for Gamma distributions is the log link function this
keeps the values positive (Kurui, 2025). For the GLM to remain
consistent for our comparison price was also used as our response
variable and maintained the same covariates. Gamma distribution was
chosen for the response variable because price is positive and
continuous and the log link function was also used again to ensure the
values remained positive as required under a Gamma distribution (Kurui,
2025).

Assumptions of our GLM and GAM models include price being continuous,
the data correctly represents the Uber and Lyft prices in Boston
Massachusetts, relationship is assumed to be linear, distribution is
positive continuous, measurements and observations are independent
(Harrell, 2024). However we did have differing assumptions on the
behaviour of the data as related to the GLM and GAM models themselves.
In our GAM we assumed the data would present non-linearly, and with our
GLM we assumed the data would present linearly.

To evaluate the models we plotted the residuals of both the GAM and GLM
to visualize how they fit the data and the variance and distribution of
the data points. We used the pseudo r-squared McFadden values and k-fold
cross validation to test the fit of our GLM. And for our GAM we used
r-squared to assess deviance explained by the model and EDF to review
the linearity of the model.

## Analysis

### Research Question #1

```{r}
gam_model <- bam(
  price ~ s(temp, k = 8) + s(wind, k = 4) + s(humidity, k = 4) +
          s(clouds, k = 5) + s(pressure, k = 5) + s(rain, k = 4) +
          distance + cab_type + name + source + destination,
  data    = merged_df,
  family  = Gamma(link = "log"),
  method  = "fREML",
  discrete = TRUE,
  select   = TRUE
)
 
summary(gam_model)
```

```{r fig.width=8, fig.height=8}
#Figure 21: Estimated Smooth Effects of Weather Variables on Log(Price)
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE, select = 1)
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE, select = 2)
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE, select = 3)
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE, select = 4)
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE, select = 5)
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE, select = 6)
```

```{r fig.width=8, fig.height=8}
#Figure 22: GAM Diagnostic Plots
par(mfrow = c(2, 2))
gam.check(gam_model)
par(mfrow = c(1, 1))
```

```{r}
cat("Adjusted R-squared:", summary(gam_model)$r.sq, "\n")
cat("Deviance explained:", round(summary(gam_model)$dev.expl * 100, 2), "%\n")
```

### Research Question #2

```{r}
glm_surge <- glm(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week +
                 source + destination + name,
  data   = lyft_df,
  family = binomial(link = "logit")
)
 
summary(glm_surge)
```

```{r}
anova(glm_surge, test = "Chisq")
```

```{r}
glmm_surge <- glmer(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week + name +
                 (1 | source) + (1 | destination),
  data    = lyft_df,
  family  = binomial(link = "logit"),
  nAGQ    = 0,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000))
)
 
summary(glmm_surge)
```

```{r}
# 1. Check random effect variance estimates
cat("=== Random Effect Variance Estimates ===\n")
vc <- VarCorr(glmm_surge)
print(vc)
 
# 2. Compare fixed effect estimates between GLM and GLMM
cat("\n=== Fixed Effect Comparison (Weather & Time) ===\n")
glm_coefs  <- coef(glm_surge)
glmm_coefs <- fixef(glmm_surge)
 
common_vars <- intersect(names(glm_coefs), names(glmm_coefs))
weather_time_pattern <- c("temp", "wind", "humidity", "clouds", "pressure",
                          "rain", "time_period", "day_of_week")
keep <- sapply(common_vars, function(x)
  any(sapply(weather_time_pattern, function(p) startsWith(x, p))))
compare_vars <- common_vars[keep]
 
comparison_df <- data.frame(
  Variable   = compare_vars,
  GLM_coef   = round(glm_coefs[compare_vars], 5),
  GLMM_coef  = round(glmm_coefs[compare_vars], 5),
  Difference = round(glm_coefs[compare_vars] - glmm_coefs[compare_vars], 5)
)
print(comparison_df, row.names = FALSE)
 
# 3. Contextual argument
cat("\n=== Contextual Justification ===\n")
cat("Number of source locations:", length(unique(lyft_df$source)), "\n")
cat("Number of destination locations:", length(unique(lyft_df$destination)), "\n")
cat("Average rides per source:", 
    round(nrow(lyft_df) / length(unique(lyft_df$source))), "\n")
cat("Average rides per destination:", 
    round(nrow(lyft_df) / length(unique(lyft_df$destination))), "\n")
```

```{r}
# McFadden pseudo R-squared for the logistic GLM
cat("McFadden Pseudo R-squared:", PseudoR2(glm_surge, which = "McFadden"), "\n")
```

```{r fig.width=8, fig.height=8}
#Figure 23: Odds Ratios for Surge Occurrence (GLM)
fe <- coef(glm_surge)
se <- summary(glm_surge)$coefficients[, "Std. Error"]
 
keep_pattern <- c("temp", "wind", "humidity", "clouds", "pressure", "rain", "time_period")
idx <- sapply(names(fe), function(x) any(sapply(keep_pattern, function(p) startsWith(x, p))))
 
or_df <- data.frame(
  Variable = names(fe)[idx],
  OR       = exp(fe[idx]),
  Lower    = exp(fe[idx] - 1.96 * se[idx]),
  Upper    = exp(fe[idx] + 1.96 * se[idx])
)
or_df$Variable <- gsub("time_period", "Time: ", or_df$Variable)
 
ggplot(or_df, aes(x = reorder(Variable, OR), y = OR)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title = "Odds Ratios for Surge Occurrence (GLM)",
       x = NULL, y = "Odds Ratio (95% CI)") +
  theme_minimal()
```

```{r fig.width=8, fig.height=8}
#Figure 24: Binned Residual Plot (GLM)
fitted_vals <- fitted(glm_surge)
resids      <- residuals(glm_surge, type = "response")
 
n_bins <- 40
bin_breaks <- quantile(fitted_vals, probs = seq(0, 1, length.out = n_bins + 1))
bins <- cut(fitted_vals, breaks = bin_breaks, include.lowest = TRUE)
 
bin_df <- data.frame(fitted_vals, resids, bins) %>%
  group_by(bins) %>%
  summarise(
    avg_fitted = mean(fitted_vals),
    avg_resid  = mean(resids),
    se_resid   = sd(resids) / sqrt(n()),
    n          = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(bins))
 
ggplot(bin_df, aes(x = avg_fitted, y = avg_resid)) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = -2 * se_resid, ymax = 2 * se_resid),
              alpha = 0.15, fill = "grey50") +
  labs(title = "Binned Residual Plot (GLM)",
       x = "Average Fitted Probability", y = "Average Residual") +
  theme_minimal()
```

### Research Question #3

```{r fig.width=8, fig.height=8}
#Figure 25: Residuals Plots - Modelling GAM
gam_model_price <- gam(price ~ holiday + day_of_the_week + s(hour) + 
                         cab_combo + surge_multiplier,
                      data = merged_data,
                      family = Gamma(link = "log"))
summary(gam_model_price)
appraise(gam_model_price)
draw(gam_model_price)
```

```{r fig.width=8, fig.height=8}
# Figure 26: Residuals Plots - Modelling GLM
glm_model_price <- glm(price ~ holiday + day_of_the_week + hour + 
                         cab_combo + surge_multiplier,
                      data = merged_data,
                      family = Gamma(link = "log"))
summary(glm_model_price)
appraise(glm_model_price)
PseudoR2(glm_model_price, which = NULL)
```

```{r}
#Validating GLM
set.seed(42)
train_control <- trainControl(method = "cv", number = 10)

glm_kfold <- train(price ~ holiday + day_of_the_week + hour + 
                     cab_combo + surge_multiplier,
                      data = merged_data,
                      method="glm",
                      family = Gamma(link = "log"),
                      trControl = train_control)
print(glm_kfold)
```

## Results

**Research Question 1:**

The GAM explained approximately 92 percent of the deviance in ride
prices, with an adjusted R squared of approximately 0.90. The dominant
predictors were distance, cab type, and service name rather than weather
conditions. The model was fit with the select equals TRUE option in bam,
which adds an extra shrinkage penalty to each smooth term. This penalty
allows the model to effectively remove smooths that do not contribute
meaningfully, providing automatic variable selection on the weather
terms rather than relying on manual inspection of p values.

**Figure 21: Estimated Smooth Effects of Weather Variables on
Log(Price):**

Figure 21 displays the estimated smooth effects for each of the six
weather variables on the log price scale, with dashed lines representing
95 percent confidence intervals and rug marks showing the distribution
of observed values. Among the weather smooth terms, cloud cover reached
high statistical significance with a p value of 0.0009 and an effective
degrees of freedom of approximately 1.89, indicating a modest non-linear
relationship with price. Atmospheric pressure was also statistically
significant with a p value of 0.017 and an effective degrees of freedom
of approximately 1.59. The remaining four weather variables were
effectively shrunk to zero by the selection penalty. Temperature had an
effective degrees of freedom of 0.05 with a p value of 0.42, while wind,
humidity, and rain all had effective degrees of freedom below 0.1 and
non-significant p values. This confirms that only cloud cover and
pressure contribute to the model, while the other weather variables have
no detectable influence on price after controlling for structural
covariates. The practical magnitude of even the two significant effects
was small. The smooth curves varied by less than 0.005 on the log price
scale, translating to price differences of roughly 5 percent or less
across the observed range of each variable.

**Figure 22: GAM Diagnostic Plots:**

Figure 22 presents the standard GAM diagnostic checks. The QQ plot of
deviance residuals in the top left panel was approximately linear,
indicating that the Gamma distribution assumption was reasonable. The
residuals versus linear predictor plot in the top right panel showed no
strong systematic patterns, suggesting that the model adequately
captured the relationship between covariates and price. The histogram of
residuals in the bottom left panel confirmed an approximately symmetric
distribution centred near zero. The response versus fitted plot in the
bottom right panel demonstrated that the model captured the general
price structure well, with points clustering around the diagonal.
Overall, the Gamma family handled the skewed distribution appropriately
and the diagnostics did not reveal major model misspecification.

In summary, only cloud cover and atmospheric pressure showed
statistically significant effects on ride prices among the six weather
variables, and the select equals TRUE option in bam confirmed that the
other four variables could be effectively removed from the model. Even
the two significant effects were practically negligible compared to
structural factors such as distance, service type, and route. This
suggests that during the observed period, the pricing algorithms of both
Uber and Lyft were not meaningfully influenced by weather conditions.

**Research Question 2:**

The McFadden pseudo R squared for the logistic GLM was 0.045. While this
may appear low, pseudo R squared values for logistic regression are not
directly comparable to OLS R squared. McFadden (1979) noted that values
between 0.2 and 0.4 represent excellent fit in discrete choice models. A
value of 0.045 is common for models predicting rare events, as the
baseline rate of surge occurrence was only 8.2 percent. The model's
value lies in identifying which predictors are associated with surge
rather than in predicting individual ride outcomes.

The analysis of deviance showed that pickup location was by far the
strongest predictor of Lyft surge pricing, contributing a deviance
reduction greater than all other predictors combined. Destination
contributed a deviance reduction of approximately 32, and time period
contributed approximately 12.5. Among the weather variables, only
atmospheric pressure reached statistical significance with a deviance
reduction of approximately 6, though even this effect was small in
practical terms.

One notable finding involved service type. After excluding Shared rides,
the service type variable contributed exactly zero deviance to the
model. This means that Lyft applies the surge multiplier uniformly
across all remaining service tiers. Whether a rider orders a regular
Lyft, Lyft XL, Lux, or Lux Black, the same surge multiplier applies.
Surge is therefore a platform level decision based on location and time
rather than a service level decision.

**Figure 23: Odds Ratios for Surge Occurrence (GLM):**

Figure \# displays the odds ratios with 95 percent confidence intervals
for the weather and time fixed effects from the GLM. Values greater than
1.0 indicate higher surge probability while the dashed line at 1.0
represents no effect. The odds ratios for weather variables all
clustered tightly around 1.0, confirming negligible effects. Morning
rush had an odds ratio of approximately 1.08 compared to midday, meaning
surge odds were about 8 percent higher during morning commute hours.
Evening rush and night showed similar but slightly smaller elevations.
Rain had an odds ratio near 1.0 but with an extremely wide confidence
interval because the vast majority of observations had zero rainfall.
The model simply did not have enough rainy rides to estimate the effect
of rain with any precision.

**Figure 24: Binned Residual Plot (GLM):**

Figure \# shows the binned residual plot for the logistic GLM, which is
the standard diagnostic for binary response models. The plot divides the
fitted probabilities into 40 bins and displays the average residual
within each bin, with grey bands indicating approximate 95 percent
confidence bounds. The majority of points fell within the expected
bounds, indicating that the model fit was adequate overall. There were a
few points near the higher fitted probability range that approached or
slightly exceeded the bounds, suggesting the model may slightly
underperform for rides with high predicted surge probability. However,
the overall pattern did not indicate serious model misspecification.

For the GLMM robustness check, the random intercept for source had a
standard deviation of 0.73 on the logit scale, confirming that
substantial variation exists between pickup locations in terms of surge
probability. This is consistent with the GLM finding that pickup
location is the dominant predictor. The destination random intercept was
near zero with a standard deviation of 0.04, indicating that the fixed
effect dummy variables already fully capture destination level
variation.

The critical result of the robustness check is the consistency of the
fixed effect estimates across the two models. All weather and time
coefficients differed by less than 0.001 on the logit scale between the
GLM and GLMM. This confirms that regardless of whether location is
treated as a fixed or random effect, the conclusions about weather and
time are virtually identical. With only 12 unique locations per grouping
variable and over 21,000 observations per group, fixed effects can
reliably estimate each location's contribution without requiring the
shrinkage that random effects provide (Bolker et al., 2009; Gelman and
Hill, 2007).

These results did not support our hypothesis. Surge pricing is not
primarily driven by temporal factors or weather. Instead, it is
dominated by pickup location. The practical implication is that a
rider's chances of encountering surge depend far more on where they ride
than on when they ride or what the weather is doing.

**Research Question 3:**

**Data Summary Results GAM:**

The data summary of the GAM indicates that cab combo is the most
significant covariate alongside surge multiplier. The EDF for the GAM is
2.5 which does indicate a degree of non-linearity, the baseline for
linear being one. Our r-squared values also indicate that the GAM
accounts for 81.6% of the deviance apparent in the model.

**Figure 25: Residuals Plots:**

The residual plots for our GAM give helpful insights into the model
behaviour. The QQ Plot indicates that the model struggles at higher
price values, this can be seen through the plot fitting the line well at
the lower and middle values but curving upwards as the values increase.
The residuals vs. Linear Predictor Plot and the Observed vs. Fitted
Values Plot both have extreme banding. This we predict is due to how
influential cab combo is on the price of Uber and Lyft ride prices. The
bands on the Linear Predictor Plot do hover around zero however they
peak in certain areas which indicates that the model is underestimating
on certain high price rides. The observed vs. Fitted Values plot
indicates there is variance being missed by the model. As described
above the model accounts for 81.6% of the variance but there is still
some that is not accounted for in this analysis. These gaps could be
from covariates which were not included in the model for this research
question. The Histogram of the Residuals is slightly right skewed as
well which indicates that the mean is higher than the median, leading us
to the conclusion we also drew from the QQ Plot which is that the model
struggles predicting at higher price points.

**Data Summary Results GLM:**

Our GLM summary indicated that cab combo and surge multiplier are the
most significant covariates for this model. The McFadden r-squared value
we calculated is a 0.2469 which according to Stack Exchange applications
of this function indicate a strong model fit (Reichenbach, 2014).

**Figure 26: Residuals Plots:**

The residual plots for our GLM are very similar to the residual plots of
our GAM. Since there was not a significant difference in the model
construction and there was only one continuous variable that was
smoothed the plots are nearly identical. Which means the model for GLM
also struggles as prices of rides increase as can be seen in the upwards
curve of the QQ Plot. There are large variances in the Residuals vs.
Linear Predictor Plot which indicate the model is underestimating the
values of prices. The Observed vs. Fitted Values Plot indicates that
there is quite a variance between the observed and predicted values.
Since the GLM model also doesn’t account for all of the deviance it is
likely that there are influences from other covariates in our dataset
that are not taken into account for this question in particular.
Finally, our histogram for the GLM is also positive right-skewed which
indicates that the model is underfitting on the higher prices which is
another confirmation for our observations above.

**K-Fold Cross Validation:**

To confirm this model we used k-fold cross validation with 10 folds.
This was to further confirm the validity of the model by testing it
against itself through withholding data. Withholding data for cross
validation testing helps us maximize our dataset by training the model
on some of our data and withholding the rest for validating the model
performance on unseen data. The outcome of our k-fold cross validation
was an RMSE of 3.98, r-squared of 81.75%, and an MAE of 2.95. The RMSE
indicates that our model observed and fitted values were off on average
by 3.98 dollars. Our r-squared score was 81.75%, which indicates how
well the model performed on the validation set. And our MAE of 2.95 is
the average difference between our price values (Farshad, 2024).

**Choosing GLM:**

We used both models to analyse the data prior to deciding on a singular
model, but ultimately we chose the GLM. Prior to editing our analysis we
had compared the GLM and GAM using AIC, which was not an accurate
comparison since different models are not directly comparable. After
reviewing the code and our analysis we looked at our existing code
summaries and model covariates practically. There was only one
continuous covariate in our model (hour), and although the EDF in the
GAM was 2.5 indicating non-linearity hour is not significant enough to
warrant keeping the GAM. As discussed in the course it makes sense to
choose the more simple model in situations like these when results and
performance are similar.

## Conclusion:

Our analysis of the Uber and Lyft pricing dataset by Munde led us to
draw new conclusions and confirmed parts of our initial hypotheses.

**Research Question 1:**

For our first research question we found that weather conditions have
almost no practical effect on ride prices. After controlling for
distance, cab type, and route, the GAM with the select equals TRUE
shrinkage penalty identified only two weather variables as contributing
to the model. Cloud cover was highly significant with a p value of
0.0009 and pressure was significant with a p value of 0.017. The
remaining four variables, temperature, wind, humidity, and rain, were
effectively shrunk to zero by the penalty, with effective degrees of
freedom near zero and non-significant p values. Even for cloud cover and
pressure, the practical magnitude was minimal, with the smooth curves
varying by less than 0.005 on the log price scale. This is the opposite
of what we predicted in our hypothesis, which was that adverse weather
would lead to meaningful price increases. The data suggest that during
this observation period, weather did not play a substantial role in the
pricing algorithms of either platform.

**Research Question 2:**

Our findings for our second research question indicated that surge
pricing is predominantly driven by pickup location, with time of day
playing a secondary role. Weather conditions showed no meaningful
association with surge occurrence. We initially hypothesized that surge
would be primarily influenced by weather and rush hour patterns rather
than by location alone, but the data clearly indicated otherwise. The
GLMM robustness check confirmed these findings. Although the source
random effect captured non-trivial between location variance, all
weather and time coefficients remained virtually identical across the
GLM and GLMM. This stability demonstrates that our conclusions are
robust to the choice of how location is modelled. With only 12 locations
and over 21,000 observations per group, fixed effects provide reliable
estimates without requiring random effects.

**Research Question 3:**

Our third and final analysis led us to the conclusion that cab combo and
surge multiplier are the primary influences on ride pricing for both
Uber and Lyft. Surge multiplier was analysed more deeply in our second
research question but in question three we did determine its strong
significance. This does align with part of our initial hypothesis where
we hypothesized cab combo would be the primary influence on price.
However, in our hypothesis we had also predicted that rush hour, holiday
and day of the week would be significant influences on price but that
was not what our findings returned. The GLM was successful in
representing the analysis of our dataset as can be confirmed by our EDA
and our model validation through k-fold cross validation which yielded
an r-squared value of 81.75%.

## Limitations:

Several limitations should be considered when interpreting these
results. The most significant is the narrow observation window. The
dataset covers only 23 days from late November to mid December 2018,
with temperatures ranging from approximately 20 to 55 degrees
Fahrenheit. This window does not include extreme weather events such as
blizzards, ice storms, or heat waves. It is plausible that weather
effects on price and surge would emerge under more extreme conditions
that our data does not contain.

The absence of Uber surge data is another important limitation. Because
the API always returned a surge multiplier of 1.0 for Uber rides, we
cannot make any claims about Uber's surge behaviour. Our surge findings
apply only to Lyft. The binary coding of surge discards information
about surge magnitude. A ride with a 1.25 multiplier and a ride with a
3.0 multiplier are treated identically. An ordinal or continuous model
could in principle recover this information, but the extreme rarity of
high surge levels in our data made such models infeasible.

The rain variable presents a particular challenge. With 59 percent of
missing values replaced by zero, our estimate of rain's effect is likely
conservative. The wide confidence interval on rain's odds ratio in the
surge model reflects this uncertainty. Rides in this dataset are not
fully independent. The same routes are sampled repeatedly across the 23
day period, and rides at nearby times likely share similar conditions.
Neither the GAM nor the GLM explicitly accounts for this temporal or
spatial correlation. While this is unlikely to change the direction of
our findings, it may affect the precision of the standard errors.

Since the timeframe for the data collection was limited to only 23 days,
we believe assumptions may have been made about data points in the
construction of the dataset. Something that indicates this limitation to
us is the fact that all cab combo categories have very similar amounts
of data. If this was simulated or the data was filtered out in
construction of the dataset it could have an impact on our final
conclusions.

The dataset locations are limited to areas of Boston Massachusetts which
limits the broader application of our conclusions. Since Uber and Lyft
are used in many regions there may be different influences on pricing
and surge multiplier in those regions that we were not able to analyse.

Looking ahead, as more data is made public on Uber and Lyft ride hailing
services in different cities and at different times of the year we
believe that it would be interesting to further analyse how these
factors influence ride-hailing pricing.

We would also like to acknowledge that due to time constraints and the
scope of the project requirements our analysis may not cover all
insights obtainable from this dataset.

## References

Anthropic. (2026). Claude (claude-sonnet-4-6) [Large language model].
<https://claude.ai>

Atlassian. (n.d.). A complete guide to violin plots. Atlassian.
<https://www.atlassian.com/data/charts/violin-plot-complete-guide>

Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear
mixed-effects models using lme4. Journal of Statistical Software, 67(1),
1–48. <https://doi.org/10.18637/jss.v067.i01>

Bolker, B. M., Brooks, M. E., Clark, C. J., Geange, S. W., Poulsen, J.
R., Stevens, M. H. H., & White, J.-S. S. (2009). Generalized linear
mixed models: a practical guide for ecology and evolution. Trends in
Ecology & Evolution, 24(3), 127–135.
<https://doi.org/10.1016/j.tree.2008.10.008>

Castillo, J. C., Knoepfle, D., & Weyl, G. (2017). Surge Pricing Solves
the Wild Goose Chase. Proceedings of the 2017 ACM Conference on
Economics and Computation. <https://doi.org/10.1145/3033274.3085098>

CRAN. (n.d.). Getting started with gratia. CRAN.
<https://cran.r-project.org/web/packages/gratia/vignettes/gratia.html>

Farshad, K. (2024). Essential Regression Evaluation Metrics: MSE, RMSE,
MAE, R², and Adjusted R². Medium.
<https://farshadabdulazeez.medium.com/essential-regression-evaluation-metrics-mse-rmse-mae-r%C2%B2-and-adjusted-r%C2%B2-0600daa1c03a>

Finnstats. (2021). Adding text labels to ggplot2 bar chart. R-Bloggers.
<https://www.r-bloggers.com/2021/09/adding-text-labels-to-ggplot2-bar-chart-2/>

Gelman, A. and Hill, J. (2007). Data Analysis Using Regression and
Multilevel/Hierarchical Models. Cambridge University Press.

Harrell, F. (2024). What Does a Statistical Method Assume? Statistical
Thinking. <https://www.fharrell.com/post/assume/>

Huang, H. (2023). Ridesharing Price Prediction: Exploring the strategies
of Dynamic Pricing. Highlights in Business Economics and Management, 10,
106–110. <https://doi.org/10.54097/hbem.v10i.7963>

Jones, M. A., Guarinello, M., Soderberg, C., & Wasser, L. A. (2020).
Time series 02: Dealing with dates & times in R - as.Date, POSIXct,
POSIXlt. National Ecological Observatory Network.
<https://www.neonscience.org/resources/learning-hub/tutorials/dc-convert-date-time-posix-r>

Kurui, M. (2025). Mastering Gamma GLM in Statsmodels: A Python Tutorial.
CodePointtech.
<https://codepointtech.com/mastering-gamma-glm-in-statsmodels-a-python-tutorial/>

Munde, R. (2019). Uber & Lyft cab prices. Kaggle.
<https://www.kaggle.com/datasets/ravi72munde/uber-lyft-cab-prices>

OpenAI. (2026). ChatGPT [Large language model]. <https://chatgpt.com/>

Osborn, L. (n.d.). Boston Weather in December. Current Results Weather
and Science Facts.
<https://www.currentresults.com/Weather/Massachusetts/Places/boston-weather-in-december.php>

Psychology Town. (2025). Exploring the concept of dispersion in data
analysis. Psychology Town.
<https://psychology.town/statistics/dispersion-concept-data-analysis/>

Reichenbach, M. (2014). McFadden's pseudo-R2 interpretation. Cross
Validated.
<https://stats.stackexchange.com/questions/82105/mcfaddens-pseudo-r2-interpretation>

RDocumentation. (2019). grep: Pattern matching and replacement.
RDocumentation.
<https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/grep>

Signorell, A., Mainwaring, B., & Wollschlaeger, D. (n.d.). Pseudo R2
statistics. CRAN.
<https://search.r-project.org/CRAN/refmans/DescTools/html/PseudoR2.html>

Signorell, A. (2025). Tools for descriptive statistics and exploratory
data analysis. RDocumentation.
<https://www.rdocumentation.org/packages/DescTools/versions/0.99.60>

Simpson, G. L., (2024). gratia: An R package for exploring generalized
additive models. Journal of Open Source Software, 9(104), 6962,
<https://doi.org/10.21105/joss.06962>

The R Graph Gallery. (n.d.). Basic barplot with ggplot2. The R Graph
Gallery.
<https://r-graph-gallery.com/218-basic-barplots-with-ggplot2.html>

The R Graph Gallery. (n.d.). ggplot2 heatmap. The R Graph Gallery.
<https://r-graph-gallery.com/79-levelplot-with-ggplot2.html>

TidyVerse. (n.d.). Violin plot. ggplot2.
<https://ggplot2.tidyverse.org/reference/geom_violin.html>

Tran, T.-A. (2025). Model diagnostics for GAM (generalised additive
model). Cross Validated.
<https://stats.stackexchange.com/questions/664508/model-diagnostics-for-gam-generalised-additive-model>

Wood, S. N. (2017). Generalized Additive Models: An Introduction with R
(2nd ed.). Chapman and Hall/CRC.

Vaisala contributors. (n.d.). Relative humidity – what is it and why is
it important? Vaisala.
<https://www.vaisala.com/en/blog/2018-10/relative-humidity-what-it-and-why-it-important>
Wikipedia Contributors. (2026). Imperial Units. Wikipedia.
<https://en.wikipedia.org/wiki/Imperial_units>

Wikipedia contributors. (2025). Cloud cover. Wikipedia.
<https://en.wikipedia.org/wiki/Cloud_cover>

## AI Use Statement

AI tools such as Claude AI, ChatGPT and Grammarly were used. For the
report, it was used to assist and enhance the writing. For the code it
was used to assist and debug the code. The content in this project and
the conclusions drawn from the data are our own.
