---
title: "surge_multiplier"
output: html_document
date: "2026-03-09"
---

# Factors Associated with Lyft Surge Pricing

## Research Question

**RQ2:** For Lyft rides, what factors, including weather conditions, time of day, day
of week, and route, are associated with the occurrence of surge pricing?


## Data Preparation

```{r surge-prep}
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(lubridate)

merged_df <- read.csv("../merged_data_cleaned.csv")

# Handle missing values
merged_df$rain[is.na(merged_df$rain)] <- 0
merged_df <- merged_df %>% filter(!is.na(price))

# Create derived variables
# Convert to Boston local time (EST = UTC-5 in Nov-Dec 2018) before extracting hour/day
merged_df <- merged_df %>%
  mutate(
    time_stamp   = as.POSIXct(time_stamp, tz = "UTC"),
    time_local   = with_tz(time_stamp, tzone = "America/New_York"),
    hour_of_day  = hour(time_local),
    # Bin hours into meaningful time periods to capture non-linear hourly patterns
    time_period  = factor(case_when(
      hour_of_day >= 6  & hour_of_day < 10 ~ "morning_rush",
      hour_of_day >= 10 & hour_of_day < 16 ~ "midday",
      hour_of_day >= 16 & hour_of_day < 20 ~ "evening_rush",
      TRUE                                  ~ "night"
    ), levels = c("midday", "morning_rush", "evening_rush", "night")),
    # Use unordered factor so R applies treatment (dummy) coding, not polynomial contrasts
    day_of_week  = factor(lubridate::wday(time_local, label = TRUE), ordered = FALSE),
    route        = paste(source, "->", destination),
    surge_binary = ifelse(surge_multiplier > 1, 1, 0)
  )
```


## Exploratory Data Analysis

```{r surge-platform-compare}
# Confirm Uber surge is always 1, only Lyft has variation
merged_df %>%
  group_by(cab_type) %>%
  summarise(
    n_rides      = n(),
    unique_surge = n_distinct(surge_multiplier),
    min_surge    = min(surge_multiplier),
    max_surge    = max(surge_multiplier),
    pct_surge    = round(mean(surge_multiplier > 1) * 100, 2),
    .groups = "drop"
  )
```

```{r surge-lyft-subset}
# Focus on Lyft only, exclude Shared rides
# Shared rides have surge_multiplier = 1 in nearly all cases, causing
# quasi-complete separation in logistic regression (coefficient -> -Inf)
lyft_df <- merged_df %>%
  filter(cab_type == "Lyft", name != "Shared")

cat("Lyft rides (excl. Shared):", nrow(lyft_df), "\n")
cat("Surge rate:", round(mean(lyft_df$surge_binary) * 100, 2), "%\n\n")
cat("Surge multiplier distribution:\n")
table(lyft_df$surge_multiplier)
```

```{r surge-by-source, fig.cap="Figure 1: Surge occurrence rate by pickup location (Lyft). Locations ordered by surge rate."}
source_surge <- lyft_df %>%
  group_by(source) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(source_surge, aes(x = reorder(source, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Pickup Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

```{r surge-by-dest, fig.cap="Figure 2: Surge occurrence rate by dropoff location (Lyft)."}
dest_surge <- lyft_df %>%
  group_by(destination) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(dest_surge, aes(x = reorder(destination, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Dropoff Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

```{r surge-by-hour, fig.cap="Figure 3: Hourly surge rate for Lyft. Higher rates during peak demand periods suggest time-of-day is an important predictor."}
hour_surge <- lyft_df %>%
  group_by(hour_of_day) %>%
  summarise(surge_rate = mean(surge_binary), n_rides = n(), .groups = "drop")

ggplot(hour_surge, aes(x = hour_of_day, y = surge_rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Hourly Surge Rate (Lyft)",
       x = "Hour of Day", y = "Surge Rate") +
  theme_minimal()
```

```{r surge-weather-compare, fig.cap="Figure 4: Standardized weather variable means for surge vs. non-surge Lyft rides. Small differences indicate weather is not the dominant surge trigger."}
weather_vars <- c("temp", "wind", "humidity", "clouds", "pressure", "rain")

# Standardize weather variables for comparable scale
lyft_std <- lyft_df %>%
  mutate(across(all_of(weather_vars), ~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE))) %>%
  mutate(Surge = ifelse(surge_binary == 1, "Surge", "No Surge")) %>%
  select(Surge, all_of(weather_vars)) %>%
  pivot_longer(-Surge, names_to = "Variable", values_to = "Value")

lyft_means <- lyft_std %>%
  group_by(Surge, Variable) %>%
  summarise(mean_val = mean(Value, na.rm = TRUE), .groups = "drop")

ggplot(lyft_means, aes(x = Variable, y = mean_val, fill = Surge)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7),
           width = 0.6, alpha = 0.8) +
  scale_fill_manual(values = c("No Surge" = "grey60", "Surge" = "steelblue")) +
  labs(title = "Weather Conditions: Surge vs. Non-Surge (Lyft)",
       subtitle = "Standardized means (z-scores); small differences suggest weather is not the primary surge driver",
       x = NULL, y = "Standardized Mean") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r surge-time-heatmap, fig.cap="Figure 5: Surge rate by time period and day of week (Lyft). Darker cells indicate higher surge probability."}
time_day_surge <- lyft_df %>%
  group_by(time_period, day_of_week) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  )

ggplot(time_day_surge, aes(x = day_of_week, y = time_period, fill = surge_rate)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = percent) +
  labs(title = "Surge Rate by Time Period and Day of Week (Lyft)",
       x = NULL, y = NULL, fill = "Surge Rate") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```


## GLM: Logistic Regression (Primary Model)

```{r surge-glm}
glm_surge <- glm(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week +
                 source + destination + name,
  data   = lyft_df,
  family = binomial(link = "logit")
)

summary(glm_surge)
```

```{r surge-glm-deviance}
# Analysis of deviance to test overall significance of each predictor group
anova(glm_surge, test = "Chisq")
```


## GLMM: Robustness Check with Random Effects

```{r surge-glmm}
library(lme4)

glmm_surge <- glmer(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week + name +
                 (1 | source) + (1 | destination),
  data    = lyft_df,
  family  = binomial(link = "logit"),
  nAGQ    = 0,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000))
)

summary(glmm_surge)
```


## Model Comparison

```{r surge-compare}
# 1. Check random effect variance estimates
cat("Random Effect Variance Estimates:\n")
print(VarCorr(glmm_surge))
cat("\nIf variances are near zero, fixed effects already capture location variation.\n\n")

# 2. Likelihood ratio test: GLM (no random effects) vs GLMM (random intercepts)
cat("Likelihood Ratio Test:\n")
lrt_stat <- -2 * (logLik(glm_surge) - logLik(glmm_surge))
lrt_df   <- 2  # two random effect variance parameters (source + destination)
lrt_p    <- pchisq(as.numeric(lrt_stat), df = lrt_df, lower.tail = FALSE)
cat("  Chi-sq statistic:", round(as.numeric(lrt_stat), 2), "\n")
cat("  df:", lrt_df, "\n")
cat("  p-value:", format.pval(lrt_p, digits = 4), "\n")
cat("  (Note: p-value is conservative due to boundary testing)\n")
```


## Odds Ratios for Fixed Effects (GLM)

```{r surge-odds-ratio, fig.cap="Figure 6: Odds ratios for weather and time fixed effects from the GLM. Error bars show 95% confidence intervals."}
fe <- coef(glm_surge)
se <- summary(glm_surge)$coefficients[, "Std. Error"]

# Select weather + time period variables
keep_pattern <- c("temp", "wind", "humidity", "clouds", "pressure", "rain", "time_period")
idx <- sapply(names(fe), function(x) any(sapply(keep_pattern, function(p) startsWith(x, p))))

or_df <- data.frame(
  Variable = names(fe)[idx],
  OR       = exp(fe[idx]),
  Lower    = exp(fe[idx] - 1.96 * se[idx]),
  Upper    = exp(fe[idx] + 1.96 * se[idx])
)
# Clean up variable names for display
or_df$Variable <- gsub("time_period", "Time: ", or_df$Variable)

ggplot(or_df, aes(x = reorder(Variable, OR), y = OR)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title = "Odds Ratios for Surge Occurrence (GLM)",
       subtitle = "Values > 1: higher surge probability; dashed line = no effect",
       x = NULL, y = "Odds Ratio (95% CI)") +
  theme_minimal()
```


## GLM Diagnostics

```{r surge-diagnostics, fig.cap="Figure 7: Binned residual plot for the GLM. Points outside the grey bands may indicate model misspecification."}
fitted_vals <- fitted(glm_surge)
resids      <- residuals(glm_surge, type = "response")

n_bins <- 40
bin_breaks <- quantile(fitted_vals, probs = seq(0, 1, length.out = n_bins + 1))
bins <- cut(fitted_vals, breaks = bin_breaks, include.lowest = TRUE)

bin_df <- data.frame(fitted_vals, resids, bins) %>%
  group_by(bins) %>%
  summarise(
    avg_fitted = mean(fitted_vals),
    avg_resid  = mean(resids),
    se_resid   = sd(resids) / sqrt(n()),
    n          = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(bins))

ggplot(bin_df, aes(x = avg_fitted, y = avg_resid)) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = -2 * se_resid, ymax = 2 * se_resid),
              alpha = 0.15, fill = "grey50") +
  labs(title = "Binned Residual Plot (GLM)",
       x = "Average Fitted Probability", y = "Average Residual") +
  theme_minimal()
```
