---
title: "surge_multiplier"
output: html_document
date: "2026-03-09"
---

# Factors Associated with Lyft Surge Pricing

## Research Question

**RQ2:** For Lyft rides, what factors, including weather conditions, time of day, day
of week, and route, are associated with the occurrence of surge pricing?


## Data Preparation

```{r surge-prep}
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(lubridate)

merged_df <- read.csv("../merged_data_cleaned.csv")

# Handle missing values
merged_df$rain[is.na(merged_df$rain)] <- 0
merged_df <- merged_df %>% filter(!is.na(price))

# Create derived variables
# Convert to Boston local time (EST = UTC-5 in Nov-Dec 2018) before extracting hour/day
merged_df <- merged_df %>%
  mutate(
    time_stamp   = as.POSIXct(time_stamp, tz = "UTC"),
    time_local   = with_tz(time_stamp, tzone = "America/New_York"),
    hour_of_day  = hour(time_local),
    # Bin hours into meaningful time periods to capture non-linear hourly patterns
    time_period  = factor(case_when(
      hour_of_day >= 6  & hour_of_day < 10 ~ "morning_rush",
      hour_of_day >= 10 & hour_of_day < 16 ~ "midday",
      hour_of_day >= 16 & hour_of_day < 20 ~ "evening_rush",
      TRUE                                  ~ "night"
    ), levels = c("midday", "morning_rush", "evening_rush", "night")),
    # Use unordered factor so R applies treatment (dummy) coding, not polynomial contrasts
    day_of_week  = factor(lubridate::wday(time_local, label = TRUE), ordered = FALSE),
    route        = paste(source, "->", destination),
    surge_binary = ifelse(surge_multiplier > 1, 1, 0)
  )
```


## Exploratory Data Analysis

```{r surge-platform-compare}
# Confirm Uber surge is always 1, only Lyft has variation
merged_df %>%
  group_by(cab_type) %>%
  summarise(
    n_rides      = n(),
    unique_surge = n_distinct(surge_multiplier),
    min_surge    = min(surge_multiplier),
    max_surge    = max(surge_multiplier),
    pct_surge    = round(mean(surge_multiplier > 1) * 100, 2),
    .groups = "drop"
  )
```

```{r surge-lyft-subset}
# Focus on Lyft only, exclude Shared rides
# Shared rides have surge_multiplier = 1 in nearly all cases, causing
# quasi-complete separation in logistic regression (coefficient -> -Inf)
lyft_df <- merged_df %>%
  filter(cab_type == "Lyft", name != "Shared")

cat("Lyft rides (excl. Shared):", nrow(lyft_df), "\n")
cat("Surge rate:", round(mean(lyft_df$surge_binary) * 100, 2), "%\n\n")
cat("Surge multiplier distribution:\n")
table(lyft_df$surge_multiplier)
```

```{r surge-by-source, fig.cap="Figure 1: Surge occurrence rate by pickup location (Lyft). Locations ordered by surge rate."}
source_surge <- lyft_df %>%
  group_by(source) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(source_surge, aes(x = reorder(source, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Pickup Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

```{r surge-by-dest, fig.cap="Figure 2: Surge occurrence rate by dropoff location (Lyft)."}
dest_surge <- lyft_df %>%
  group_by(destination) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(surge_rate))

ggplot(dest_surge, aes(x = reorder(destination, surge_rate), y = surge_rate)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Surge Rate by Dropoff Location (Lyft)",
       subtitle = "Percentage of rides with surge_multiplier > 1",
       x = NULL, y = "Surge Rate") +
  theme_minimal()
```

```{r surge-by-hour, fig.cap="Figure 3: Hourly surge rate for Lyft. Higher rates during peak demand periods suggest time-of-day is an important predictor."}
hour_surge <- lyft_df %>%
  group_by(hour_of_day) %>%
  summarise(surge_rate = mean(surge_binary), n_rides = n(), .groups = "drop")

ggplot(hour_surge, aes(x = hour_of_day, y = surge_rate)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_y_continuous(labels = percent) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Hourly Surge Rate (Lyft)",
       x = "Hour of Day", y = "Surge Rate") +
  theme_minimal()
```

```{r surge-weather-compare, fig.cap="Figure 4: Standardized weather variable means for surge vs. non-surge Lyft rides. Small differences indicate weather is not the dominant surge trigger."}
weather_vars <- c("temp", "wind", "humidity", "clouds", "pressure", "rain")

# Standardize weather variables for comparable scale
lyft_std <- lyft_df %>%
  mutate(across(all_of(weather_vars), ~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE))) %>%
  mutate(Surge = ifelse(surge_binary == 1, "Surge", "No Surge")) %>%
  select(Surge, all_of(weather_vars)) %>%
  pivot_longer(-Surge, names_to = "Variable", values_to = "Value")

lyft_means <- lyft_std %>%
  group_by(Surge, Variable) %>%
  summarise(mean_val = mean(Value, na.rm = TRUE), .groups = "drop")

ggplot(lyft_means, aes(x = Variable, y = mean_val, fill = Surge)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7),
           width = 0.6, alpha = 0.8) +
  scale_fill_manual(values = c("No Surge" = "grey60", "Surge" = "steelblue")) +
  labs(title = "Weather Conditions: Surge vs. Non-Surge (Lyft)",
       subtitle = "Standardized means (z-scores); small differences suggest weather is not the primary surge driver",
       x = NULL, y = "Standardized Mean") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r surge-time-heatmap, fig.cap="Figure 5: Surge rate by time period and day of week (Lyft). Darker cells indicate higher surge probability."}
time_day_surge <- lyft_df %>%
  group_by(time_period, day_of_week) %>%
  summarise(
    surge_rate = mean(surge_binary),
    n_rides    = n(),
    .groups = "drop"
  )

ggplot(time_day_surge, aes(x = day_of_week, y = time_period, fill = surge_rate)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = percent(surge_rate, accuracy = 0.1)), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue", labels = percent) +
  labs(title = "Surge Rate by Time Period and Day of Week (Lyft)",
       x = NULL, y = NULL, fill = "Surge Rate") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```


## GLM: Logistic Regression (Primary Model)

```{r surge-glm}
glm_surge <- glm(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week +
                 source + destination + name,
  data   = lyft_df,
  family = binomial(link = "logit")
)

summary(glm_surge)
```

```{r surge-glm-deviance}
# Analysis of deviance to test overall significance of each predictor group
anova(glm_surge, test = "Chisq")
```


## GLMM: Robustness Check with Random Effects

```{r surge-glmm}
library(lme4)

glmm_surge <- glmer(
  surge_binary ~ temp + wind + humidity + clouds + pressure + rain +
                 time_period + day_of_week + name +
                 (1 | source) + (1 | destination),
  data    = lyft_df,
  family  = binomial(link = "logit"),
  nAGQ    = 0,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 50000))
)

summary(glmm_surge)
```


## Model Comparison: GLM vs GLMM Robustness Check

We cannot directly compare GLM and GLMM using AIC or a standard likelihood
ratio test, because the two models use different likelihood formulations:
the GLM maximizes a conditional likelihood while the GLMM integrates over
random effects using a Laplace approximation. As noted in the course
materials (Week 3, GLMM Theory), standard AIC is only appropriate for
comparing GLMM variants when we are interested in accounting for
correlation structures; when evaluating whether random effects are
necessary, AIC must be modified to conditional AIC (cAIC), which is beyond
the scope of this analysis.

Instead, we assess whether the GLMM's random effects add meaningful
information by examining: 
(1) the estimated random effect variances
(2) the consistency of fixed effect estimates between the two models
(3) the practical context of our grouping variables

```{r surge-compare}
# 1. Check random effect variance estimates
cat("=== Random Effect Variance Estimates ===\n")
vc <- VarCorr(glmm_surge)
print(vc)
cat("\nInterpretation: If variances are near zero, fixed effects already\n")
cat("capture the location-level variation adequately.\n\n")

# 2. Compare fixed effect estimates between GLM and GLMM
cat("=== Fixed Effect Comparison (Weather & Time) ===\n")
glm_coefs  <- coef(glm_surge)
glmm_coefs <- fixef(glmm_surge)

# Identify common coefficients (weather + time variables)
common_vars <- intersect(names(glm_coefs), names(glmm_coefs))
weather_time_pattern <- c("temp", "wind", "humidity", "clouds", "pressure",
                          "rain", "time_period", "day_of_week")
keep <- sapply(common_vars, function(x)
  any(sapply(weather_time_pattern, function(p) startsWith(x, p))))
compare_vars <- common_vars[keep]

comparison_df <- data.frame(
  Variable   = compare_vars,
  GLM_coef   = round(glm_coefs[compare_vars], 5),
  GLMM_coef  = round(glmm_coefs[compare_vars], 5),
  Difference = round(glm_coefs[compare_vars] - glmm_coefs[compare_vars], 5)
)
print(comparison_df, row.names = FALSE)

cat("\nIf fixed effect estimates are nearly identical between GLM and GLMM,\n")
cat("this confirms that the random intercepts do not meaningfully change\n")
cat("the substantive conclusions.\n\n")

# 3. Contextual argument
cat("=== Contextual Justification ===\n")
cat("Number of source locations:", length(unique(lyft_df$source)), "\n")
cat("Number of destination locations:", length(unique(lyft_df$destination)), "\n")
cat("Average rides per source:", round(nrow(lyft_df) / length(unique(lyft_df$source))), "\n")
cat("Average rides per destination:", round(nrow(lyft_df) / length(unique(lyft_df$destination))), "\n")
cat("\nWith only 12 unique locations per grouping variable and thousands of\n")
cat("observations per group, fixed effects can fully capture location-level\n")
cat("variation. Random effects are most beneficial when there are many groups\n")
cat("with few observations each (Bolker et al., 2009; Gelman & Hill, 2007).\n")
```


## Odds Ratios for Fixed Effects (GLM)

```{r surge-odds-ratio, fig.cap="Figure 6: Odds ratios for weather and time fixed effects from the GLM. Error bars show 95% confidence intervals."}
fe <- coef(glm_surge)
se <- summary(glm_surge)$coefficients[, "Std. Error"]

# Select weather + time period variables
keep_pattern <- c("temp", "wind", "humidity", "clouds", "pressure", "rain", "time_period")
idx <- sapply(names(fe), function(x) any(sapply(keep_pattern, function(p) startsWith(x, p))))

or_df <- data.frame(
  Variable = names(fe)[idx],
  OR       = exp(fe[idx]),
  Lower    = exp(fe[idx] - 1.96 * se[idx]),
  Upper    = exp(fe[idx] + 1.96 * se[idx])
)
# Clean up variable names for display
or_df$Variable <- gsub("time_period", "Time: ", or_df$Variable)

ggplot(or_df, aes(x = reorder(Variable, OR), y = OR)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2, color = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(title = "Odds Ratios for Surge Occurrence (GLM)",
       subtitle = "Values > 1: higher surge probability; dashed line = no effect",
       x = NULL, y = "Odds Ratio (95% CI)") +
  theme_minimal()
```


## GLM Diagnostics

```{r surge-diagnostics, fig.cap="Figure 7: Binned residual plot for the GLM. Points outside the grey bands may indicate model misspecification."}
fitted_vals <- fitted(glm_surge)
resids      <- residuals(glm_surge, type = "response")

n_bins <- 40
bin_breaks <- quantile(fitted_vals, probs = seq(0, 1, length.out = n_bins + 1))
bins <- cut(fitted_vals, breaks = bin_breaks, include.lowest = TRUE)

bin_df <- data.frame(fitted_vals, resids, bins) %>%
  group_by(bins) %>%
  summarise(
    avg_fitted = mean(fitted_vals),
    avg_resid  = mean(resids),
    se_resid   = sd(resids) / sqrt(n()),
    n          = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(bins))

ggplot(bin_df, aes(x = avg_fitted, y = avg_resid)) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = -2 * se_resid, ymax = 2 * se_resid),
              alpha = 0.15, fill = "grey50") +
  labs(title = "Binned Residual Plot (GLM)",
       x = "Average Fitted Probability", y = "Average Residual") +
  theme_minimal()
```
