---
title: "weather_effect"
output: html_document
date: "2026-03-06"
---

# Effect of Weather Conditions on Ride Price

## Research Question

**RQ1:** After controlling for distance, route, cab type, and service level, do weather
conditions (temperature, wind, humidity, cloud cover, pressure, and rain) have a
significant effect on ride price, and is the relationship non-linear?


## Data Preparation and Feature Engineering

```{r weather-prep}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)

merged_df <- read.csv("../merged_data_cleaned.csv")

# Replace rain NA with 0 (NA indicates no rain was recorded)
merged_df$rain[is.na(merged_df$rain)] <- 0

# Remove rows with missing price
merged_df <- merged_df %>% filter(!is.na(price))

# Create useful derived variables
# Data was collected in Boston (EST = UTC-5 in Nov-Dec 2018)
merged_df <- merged_df %>%
  mutate(
    time_stamp  = as.POSIXct(time_stamp, tz = "UTC"),
    hour_ts     = as.POSIXct(hour_ts, tz = "UTC"),
    time_local  = with_tz(time_stamp, tzone = "America/New_York"),
    hour_of_day = hour(time_local),
    day_of_week = lubridate::wday(time_local, label = TRUE),
    cab_label   = paste(cab_type, name, sep = "-"),
    route       = paste(source, "->", destination),
    surge_binary = ifelse(surge_multiplier > 1, 1, 0)
  )

cat("Dataset dimensions:", nrow(merged_df), "rows x", ncol(merged_df), "columns\n")
cat("Date range:", as.character(min(merged_df$time_stamp)),
    "to", as.character(max(merged_df$time_stamp)), "\n")
```


## Service Type Overview

```{r cab-label-analysis}
# Order quantity and average price for each cab_type-name combination
cab_summary <- merged_df %>%
  group_by(cab_label) %>%
  summarise(
    order_count = n(),
    avg_price   = round(mean(price, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(order_count))

print(cab_summary)
```


## Exploratory Data Analysis

```{r weather-summary}
# Summary statistics for weather variables
merged_df %>%
  select(temp, clouds, pressure, humidity, wind, rain) %>%
  summary()
```

```{r weather-price-hist, fig.cap="Figure 1: Distribution of ride prices. The strong right skew supports using a Gamma family GAM rather than Gaussian regression."}
ggplot(merged_df, aes(x = price)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Ride Prices",
       x = "Price (USD)", y = "Count") +
  theme_minimal()
```

```{r weather-temp-scatter, fig.cap="Figure 2: Temperature vs. price (log scale) by platform. A GAM smoother on a random subsample shows the trend."}
set.seed(42)
sample_df <- merged_df %>% slice_sample(n = 20000)

ggplot(merged_df, aes(x = temp, y = price)) +
  geom_point(alpha = 0.02, size = 0.3, color = "grey40") +
  geom_smooth(data = sample_df, method = "gam", formula = y ~ s(x),
              se = TRUE, color = "steelblue", linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~ cab_type) +
  labs(title = "Temperature vs. Price (Log Scale) by Platform",
       x = "Temperature (F)", y = "Price (USD, log scale)") +
  theme_minimal()
```

```{r weather-heatmap, fig.cap="Figure 3: Pearson correlation between weather variables and price for each service type. These are unadjusted and do not control for distance or route."}
cor_results <- merged_df %>%
  group_by(cab_label) %>%
  summarise(
    Temperature = cor(price, temp,     use = "complete.obs"),
    Clouds      = cor(price, clouds,   use = "complete.obs"),
    Pressure    = cor(price, pressure, use = "complete.obs"),
    Humidity    = cor(price, humidity,  use = "complete.obs"),
    Wind        = cor(price, wind,     use = "complete.obs"),
    Rain        = cor(price, rain,     use = "complete.obs"),
    .groups = "drop"
  )

cor_long <- cor_results %>%
  pivot_longer(-cab_label, names_to = "Weather Variable", values_to = "Correlation")

ggplot(cor_long, aes(x = `Weather Variable`, y = cab_label, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 3)), size = 2.8) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  labs(title = "Correlation Between Weather and Price by Service Type",
       subtitle = "Unadjusted Pearson correlations (not controlling for distance or route)",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```


## GAM Analysis

```{r gam-fit}
library(mgcv)
# bam() is the large dataset version of gam(); equivalent to gam() but optimized for large N via discretized covariate methods.
# Gamma family + log link handles right-skewed positive prices where variance increases with the mean.

# Basis dimension (k) selection rationale:
#   - temp (k=8): temperature has the widest range and most potential for complex non-linearity; higher k allows more flexibility.
#   - clouds, pressure (k=5): moderate potential for non-linear effects.
#   - wind, humidity, rain (k=4): limited variability in the data or sparse observations (rain is mostly zero); lower k avoids over fitting.
# The effective degrees of freedom (edf) reported by summary() indicate how much of this flexibility the data actually support.
gam_model <- bam(
  price ~ s(temp, k = 8) + s(wind, k = 4) + s(humidity, k = 4) +
          s(clouds, k = 5) + s(pressure, k = 5) + s(rain, k = 4) +
          distance + cab_type + name + source + destination,
  data    = merged_df,
  family  = Gamma(link = "log"),
  method  = "fREML",
  discrete = TRUE
)

summary(gam_model)
```


## Smooth Effect Plots

```{r gam-smooth-plots, fig.height=8, fig.cap="Figure 4: Estimated smooth effects of weather variables on log(price). Shaded areas are 95% CIs."}
par(mfrow = c(3, 2), mar = c(4, 4, 2, 1))
plot(gam_model, shade = TRUE, shade.col = "lightblue",
     residuals = FALSE, rug = TRUE, se = TRUE)
par(mfrow = c(1, 1))
```


## Model Diagnostics

```{r gam-diagnostics, fig.height=7, fig.cap="Figure 5: GAM diagnostic plots. Top-left: QQ plot. Top-right: residuals vs. linear predictor. Bottom-left: histogram of residuals. Bottom-right: response vs. fitted."}
par(mfrow = c(2, 2))
gam.check(gam_model)
par(mfrow = c(1, 1))
```

```{r gam-fit-stats}
cat("Adjusted R-squared:", summary(gam_model)$r.sq, "\n")
cat("Deviance explained:", round(summary(gam_model)$dev.expl * 100, 2), "%\n")
```
